---
title: "Homework Week 6: The Logic of Bayesian Inference"
subtitle: "QMIR 2026 -- Due one week from today"
author: "Your Name"
date: today
date-format: MMMM D, YYYY
format:
  pdf:
    toc: false
    number-sections: false
    echo: true
    code-fold: false
execute:
  message: false
  warning: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
```

# Setting: Alien Democracy Detectors

An alien civilisation is trying to answer a scientific question about Earth: **what proportion of land area belongs to democratic countries?**

Their method is simple but clever. They dispatch probe ships at random. Each probe avoids the oceans (there is nothing to land on) and touches down at a uniformly random location on land. Upon landing, the probe's instruments identify the country, cross-reference it with the V-Dem dataset, and record a binary outcome:

- $D = 1$ -- the country is classified as a **democracy**
- $D = 0$ -- the country is classified as a **non-democracy**

The aliens treat each landing as an independent observation. After $N$ landings, they have recorded $W$ democratic touchdowns. Their goal is to estimate $\theta$, the true proportion of land area belonging to democracies.

You have been assigned to help them reason carefully about this.

# Part I: Conceptual Questions

Answer each question in **2–4 sentences** in your own words. You do not need any code for this part.

## What is the parameter of interest?

Describe in plain language what $\theta$ represents in this scenario. Is it the same as "what fraction of countries are democracies"? Why or why not?

*Your answer here.*

## Choosing a prior

The alien scientists have three competing prior beliefs about $\theta$ before any probes land:

- **Flat prior:** complete agnosticism -- every value of $\theta$ between 0 and 1 is equally plausible.
- **Weakly informative prior:** the scientists have read some Earth political science and believe democracies are somewhat more common than non-democracies, but are uncertain. They give modestly more prior weight to values of $\theta > 0.5$.
- **Supercertain prior:** one senior alien scientist is convinced -- based on intercepted radio signals -- that exactly half of Earth's land belongs to democracies. She concentrates nearly all prior mass at $\theta = 0.5$.

For each prior, write one sentence: is this a scientifically defensible prior for this research question? What assumption does it encode?

*Your answer here.*

## The likelihood family

The aliens model each probe landing as a Bernoulli trial: with probability $\theta$, the probe lands in a democratic country; with probability $1 - \theta$, it does not. With $N$ landings, the count of democratic landings $W$ follows a Binomial distribution.

In your own words: what does the likelihood $P(W, N \mid \theta)$ tell us? Why is it not a probability distribution over $\theta$?

*Your answer here.*

## Interpreting the posterior

After the probes return their data, the alien scientists compute a posterior distribution over $\theta$.

A junior scientist summarises the result: *"The 90% credible interval for $\theta$ is [0.42, 0.71]. This means there is a 90% probability that the true proportion of democratic land lies between 0.42 and 0.71."*

Is this interpretation correct under the Bayesian framework? Would this interpretation be valid for a frequentist 90% confidence interval? Explain the difference briefly.

*Your answer here.*


# Part II: Bayesian Updating in R

## The playground

The code block below implements the full Bayesian grid approximation for the alien probe scenario. The parameters at the top are labelled clearly -- **your job is to change them and observe what happens**.

```{r}
#| label: alien-playground

# ── Change these to explore ───────────────────────────────────────────────────
W <- 4 # democratic landings observed
N <- 6 # total probe landings
prior_type <- "flat" # "flat", "weak", or "supercertain"

# ── Grid over all possible θ values ──────────────────────────────────────────
theta_grid <- seq(0, 1, length.out = 1000)

# ── Prior ─────────────────────────────────────────────────────────────────────
prior <- switch(
  prior_type,
  flat = rep(1, 1000),
  weak = ifelse(theta_grid > 0.5, 2, 1),
  supercertain = dnorm(theta_grid, mean = 0.5, sd = 0.02)
)
prior <- prior / sum(prior)

# ── Likelihood: Binomial ──────────────────────────────────────────────────────
likelihood <- dbinom(W, size = N, prob = theta_grid)

# ── Bayes' theorem ────────────────────────────────────────────────────────────
numerator <- likelihood * prior
normalizing_const <- sum(numerator) # P(data)
posterior <- numerator / normalizing_const

# To see that P(data) is just a number
cat("P(data) =", round(normalizing_const, 5), "\n")

# To see the posterior sums to 1 (i.e. is a valid probability distribution)
sum(posterior)

# ── Plot ──────────────────────────────────────────────────────────────────────
tibble(
  theta = theta_grid,
  Prior = prior / max(prior),
  Likelihood = likelihood / max(likelihood),
  Posterior = posterior / max(posterior)
) |>
  pivot_longer(-theta, names_to = "dist", values_to = "val") |>
  mutate(dist = factor(dist, levels = c("Prior", "Likelihood", "Posterior"))) |>
  ggplot(aes(x = theta, y = val, color = dist, linetype = dist)) +
  geom_line(linewidth = 0.9) +
  scale_color_manual(
    values = c("steelblue", "goldenrod3", "firebrick"),
    name = NULL
  ) +
  scale_linetype_manual(values = c("solid", "dashed", "twodash"), name = NULL) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    aspect.ratio = 0.42,
    panel.grid.minor = element_blank()
  ) +
  labs(
    x = expression(theta ~ "(proportion of democratic land)"),
    y = "Scaled density (max = 1)",
    caption = paste0("prior = ", prior_type, "  |  W = ", W, "  |  N = ", N)
  )
```

## Exploring: flat prior

Run the code with `prior_type = "flat"` and `W = 4, N = 6`.

In 2–3 sentences: what shape does the posterior have, and why does it look the way it does? What is the relationship between the posterior and the likelihood under a flat prior?

*Your answer here.*

## Exploring: the effect of sample size

Keep `prior_type = "flat"` and the proportion of democratic landings constant at $W/N = 2/3$. Run the code for three sample sizes: `W = 4, N = 6`; `W = 40, N = 60`; `W = 400, N = 600`. Include all three plots in your submission.

In 3–4 sentences: describe what changes across the three plots and what stays the same. What does this tell you about the relationship between data and uncertainty?

```{r}
#| label: sample-size-small
# W = 4, N = 6 (already run above -- copy or re-run here)

```

```{r}
#| label: sample-size-medium
# W = 40, N = 60

```

```{r}
#| label: sample-size-large
# W = 400, N = 600

```

*Your answer here.*

## Exploring: the effect of the prior

Use `W = 4, N = 6` and run the code separately for all three priors. Include all three plots.

In 3–4 sentences: how does the supercertain prior change the posterior compared to the flat prior? Does the weak prior make a visible difference? Which prior would you defend as the most scientifically appropriate starting point for this alien research question, and why?

```{r}
#| label: prior-flat
# prior_type = "flat"

```

```{r}
#| label: prior-weak
# prior_type = "weak"

```

```{r}
#| label: prior-supercertain
# prior_type = "supercertain"

```

*Your answer here.*

## Posterior summary

Using any one of your posteriors from the previous section (state which one), you draw 10,000 samples from it and compute summary statistics using the `sample` function in the code below.

```{r}
#| label: posterior-summary

# ── Re-run your chosen setup here first ──────────────────────────────────────
W <- 4
N <- 6
prior_type <- "flat" # <- change if desired

theta_grid <- seq(0, 1, length.out = 1000)
prior <- switch(
  prior_type,
  flat = rep(1, 1000),
  weak = ifelse(theta_grid > 0.5, 2, 1),
  supercertain = dnorm(theta_grid, mean = 0.5, sd = 0.02)
)
prior <- prior / sum(prior)
likelihood <- dbinom(W, size = N, prob = theta_grid)
posterior <- (likelihood * prior) / sum(likelihood * prior)

# ── Draw samples from the posterior ──────────────────────────────────────────
set.seed(42)
samples <- sample(theta_grid, size = 1e4, replace = TRUE, prob = posterior)

# ── Summarise ─────────────────────────────────────────────────────────────────
cat("Posterior mean:  ", round(mean(samples), 3), "\n")
cat("Posterior median:", round(median(samples), 3), "\n")
cat(
  "90% credible interval:",
  round(quantile(samples, 0.05), 3),
  "--",
  round(quantile(samples, 0.95), 3),
  "\n"
)

# ── Plot ──────────────────────────────────────────────────────────────────────
tibble(theta = samples) |>
  ggplot(aes(x = theta)) +
  geom_histogram(
    bins = 50,
    fill = "steelblue",
    color = "white",
    linewidth = 0.2
  ) +
  geom_vline(
    xintercept = quantile(samples, c(0.05, 0.95)),
    linetype = "dashed",
    color = "firebrick"
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(
    x = expression(theta ~ "(proportion of democratic land)"),
    y = "Count",
    caption = paste0(
      "10,000 posterior samples | prior = ",
      prior_type,
      " | W = ",
      W,
      " | N = ",
      N
    )
  )
```

Write **one sentence** interpreting the 90% credible interval printed above in plain, non-technical language -- as you would explain it to the alien scientists.

*Your answer here.*



# Submission checklist

Before submitting, make sure your rendered PDF:

- [ ] Answers all four conceptual questions in Part I
- [ ] Contains six plots from Part II (three sample sizes + three priors)
- [ ] Contains the posterior summary plot with a written interpretation
- [ ] Was produced by rendering this `.qmd` file (not copy-pasted output)
- [ ] Includes your name in the YAML header above
