---
title: "Homework 03: R fundamentals"
author: "Your name"
date: last-modified
date-format: MMMM D, YYYY
format:
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
execute:
  echo: true
  message: false
  warning: false
---

<!-- STUDENT-FACING STARTER. No solution content here. Files named solution.* are stripped on
     release, but this file ships as-is, so keep it clean.

     Base R only this week: no packages are loaded, so the document renders anywhere R does.
     Section 1 is a worked demo of code cells. The exercises start at section 2. -->

```{r}
#| label: setup
#| include: false

start_time <- Sys.time()
```

# Code cells in Quarto {#sec-code-cells}

This section is a demonstration, not an exercise.
Read it, render the document once, and compare the PDF with the source.

## What a code cell looks like {#sec-cell-anatomy}

A code cell starts with three backticks followed by `{r}`, and ends with three backticks.
Everything in between is R code.
Lines starting with `#|` are **cell options**, and they tell Quarto how to treat the cell.
Each option goes on its own `#|` line at the top of the cell.
In the source, a cell looks like this:

```{{r}}
#| label: first-cell

1 + 1
```

And this is the same cell, run for real, with its own label because labels must be unique.
When you render, Quarto runs the code and puts the result directly underneath it:

```{r}
#| label: first-cell-run

1 + 1
```

## Cell options you will use {#sec-cell-options}

| Option | What it does |
|---|---|
| `label: my-name` | Names the cell. Every cell gets a label, and labels must be unique. |
| `echo: false` | Runs the code but hides it, so only the result appears. |
| `eval: false` | Shows the code but does not run it. |
| `error: true` | Lets the render continue when the cell produces an error, and prints the error message. |

: The four code cell options this homework uses. {#tbl-cell-options}

The next cell has `eval: false`, so the code appears in the PDF but produces no output:

```{r}
#| label: not-run
#| eval: false

sqrt(16)
```

The next cell has `error: true`.
Without it, the error would stop the whole render.

```{r}
#| label: an-error
#| error: true

sqrt("sixteen")
```

## Running cells versus rendering {#sec-run-vs-render}

In Positron you can run a single cell with the **Run Cell** button above it, or run the current line with Ctrl+Enter (Cmd+Enter on a Mac).
The result appears in the console, which is how you try things out while you work.

**Rendering** is different: Quarto starts a fresh R session and runs every cell from top to bottom.

::: {.callout-important}
**Your document must work from top to bottom in a fresh session.**
An object you created only in the console does not exist when Quarto renders.
If the render fails with `object not found` while your console works fine, this is almost always the reason.
:::

## Code inside a sentence {#sec-inline-code}

You can also put a result inside your prose with inline code: a backtick, the letter `r`, a space, the code, and a closing backtick.
Written that way, `2 + 2` becomes `r 2 + 2` in the rendered document, and it stays correct if the code changes.

# Objects {#sec-objects}

1. Create three objects: one holding a number, one holding a piece of text, and one holding `TRUE` or `FALSE`.
   Choose meaningful names.
2. Print all three.
3. Assign a new value of a different type to one of them, and print it again.

```{r}
#| label: ex-objects
```

In one sentence, describe what happened to the object in step 3:

<!-- Your answer here. -->

# Functions {#sec-functions}

1. Create a vector of five numbers of your choice.
   Compute its `mean()`, `min()` and `max()`, and round the mean to two decimal places with `round()`, naming the `digits` argument.

```{r}
#| label: ex-functions-builtin
```

2. Write a function `km_to_miles()` that converts kilometres to miles (1 km is 0.621371 miles).
   Test it on a single value and on a vector of three distances.

```{r}
#| label: ex-functions-km
```

3. Write a function `percent()` that takes a proportion `x` and returns it as a percentage, rounded to `digits` decimal places.
   Give `digits` a default value of 1.
   Call it once with and once without the `digits` argument.

```{r}
#| label: ex-functions-percent
```

# Reading errors {#sec-errors}

Each line in the next cell produces an error.
The cell has `error: true`, so the document still renders and shows each message.

```{r}
#| label: ex-errors-broken
#| error: true

countrie <- c("Germany", "France")
country_count <- length(countries)
Sum(c(1, 2, 3))
"10" * 2
```

1. For each of the three error messages, write one sentence saying **where** it broke and **what** went wrong.

<!-- Your answer here. -->

2. Write a fixed version of all four lines in the next cell, which does not have `error: true`.

```{r}
#| label: ex-errors-fixed
```

# The pipe {#sec-pipe}

The next cell computes something in one nested call.

```{r}
#| label: ex-pipe-nested

distances <- c(4.2, 10, 21.1, 42.2)
round(sum(distances), digits = 0)
```

1. Describe in one sentence what the nested call computes.
2. Rewrite it with the native pipe `|>`, and show that both versions give the same result.
3. Extend the pipe so that it first converts `distances` to miles with your `km_to_miles()` function.

```{r}
#| label: ex-pipe
```

<!-- Your answer to question 1 here. -->

# Data types {#sec-types}

1. **Before running anything**, write down which class you expect for each value: `42`, `42L`, `"42"`, `FALSE`, `"FALSE"`, `10 > 5`, `as.Date("2026-09-30")`.

<!-- Your predictions here. -->

2. Now check each one with `class()`.
   Which of your predictions were wrong, and why?

```{r}
#| label: ex-types-class
```

<!-- Your answer here. -->

3. Convert `"2026"` to a number and add 1 to it.
   Then try `as.numeric("twenty")`.
   What comes back, and what does it mean?

```{r}
#| label: ex-types-convert
#| warning: true
```

<!-- Your answer here. -->

4. Create a factor `size` from the values `"small"`, `"large"`, `"medium"`, `"small"`, with the levels in the order small, medium, large.
   Print it and its `levels()`.

```{r}
#| label: ex-types-factor
```

# Data structures {#sec-structures}

## Vectors and matrices {#sec-ex-vectors}

1. Create a vector with the names of four cities, and check its `length()`.
2. Create a numeric vector with their populations in thousands, and use one line of code to convert all of them to plain numbers of inhabitants.
3. **Predict first**, then check: what does `c(3, "four", 5)` return, and what is its class?
4. Create a matrix with 2 rows and 3 columns holding the numbers 1 to 6.

```{r}
#| label: ex-vectors
```

## A small dataset {#sec-ex-dataset}

The next cell builds the data frame `europe`, which the rest of this homework uses.
Run it, but do not change it.
Populations are in millions for 2024, rounded, and `eu_joined` is the year the country joined the EU or its predecessor.

```{r}
#| label: europe-data

europe <- data.frame(
  country = c(
    "Germany", "France", "Italy", "Spain", "Poland",
    "Sweden", "Norway", "Switzerland", "Turkey"
  ),
  population = c(83.4, 68.4, 59.0, 48.6, 36.6, 10.6, 5.6, 9.0, 85.4),
  eu_member = c(TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, FALSE, FALSE, FALSE),
  eu_joined = c(1958, 1958, 1958, 1986, 2004, 1995, NA, NA, NA),
  region = c(
    "Western Europe", "Western Europe", "Southern Europe", "Southern Europe",
    "Eastern Europe", "Northern Europe", "Northern Europe", "Western Europe",
    "Western Asia"
  )
)
```

1. Look at `europe` with `str()`.
   How many rows and columns does it have, and what type is each column?

```{r}
#| label: ex-dataset-str
```

<!-- Your answer here. -->

2. `NA` appears in `eu_joined`.
   In one sentence, what does it stand for here, and why is it a better choice than 0?

<!-- Your answer here. -->

## A list {#sec-ex-list}

Create a list `me` holding your name, the number of semesters you have studied, and a vector of two courses you are taking this term.
Print it, then print only the courses.

```{r}
#| label: ex-list
```

# Accessing and exploring data {#sec-access}

All questions use `europe`.

1. Print the third row.
2. Print the `population` column as a vector, using `$`.
3. Print the names of all countries with more than 50 million inhabitants.
4. How many of the nine countries are EU members?
   Answer with code, not by counting.
5. Which countries joined after 2000?
6. Compute the average population with `summary()`.
7. Compute the average accession year with `mean()`.
   The first attempt returns `NA`.
   Use `?mean` to find the argument that fixes it.

```{r}
#| label: ex-access
```

# Reflection {#sec-reflection .unnumbered}

What did you find confusing or difficult in this homework?
Is there anything about objects, functions, types or structures you are still unsure about?

<!-- Your answer here. -->

# Session info and execution time {#sec-session .unnumbered}

```{r}
#| label: session-info

sessionInfo()
```

```{r}
#| label: exec-time

end_time <- Sys.time()
cat(
  "Total execution time:",
  round(difftime(end_time, start_time, units = "mins"), 1),
  "minutes\n"
)
```
