---
title: |
  Week 08 Homework
subtitle: |
  Bivariate Bayesian Linear Regression
author: "Your name"
date: last-modified
date-format: MMMM D, YYYY
format:
  pdf:
    toc: true
    include-in-header:
      text: |
        \usepackage{setspace}
        \setlength{\parindent}{15pt}
execute:
  echo: true
  warning: false
  message: false
number-sections: true
papersize: a4
fontsize: 12pt
linestretch: 1.5
geometry:
  - top = 2cm
  - bottom = 2cm
  - left = 2.5cm
  - right = 2.5cm
  - footskip = 20pt
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(brms)
library(bayesplot)
library(patchwork)
```

This homework walks you through the **complete bivariate Bayesian regression workflow** from Week 8, applied to a new substantive question.

**The research question**: *Does institutional quality predict how much foreign direct investment a country attracts?*

**The dataset**: `data/fdi_governance.csv` contains 180 synthetic country observations. The variables are:

| Variable | Description |
|---|---|
| `country_id` | Unique country identifier (C001–C180) |
| `region` | World region (World Bank classification) |
| `gov_effectiveness` | Government effectiveness index, standardized (mean 0, SD 1; higher = better governance) |
| `fdi_pct_gdp` | Net FDI inflows as a percentage of GDP — the **outcome variable** |

Work through the sections in order. Your rendered PDF should contain all code, plots, and written answers.

# Data Exploration

## Load and inspect

Load the dataset from `data/fdi_governance.csv`. Use `glimpse()` to confirm the variable types and number of observations.

```{r}
#| label: load-data

# Your code here

```

## Descriptive statistics

Compute the mean, standard deviation, minimum, and maximum for both `gov_effectiveness` and `fdi_pct_gdp`. You may use `summarise()` with `across()` or simply call `summary()`.

In **one sentence**, interpret the mean of `fdi_pct_gdp`.

```{r}
#| label: desc-stats

# Your code here

```

*Your interpretation:*

## Marginal distribution of the outcome

Produce a histogram of `fdi_pct_gdp`. Include informative axis labels and a title.

Does the distribution appear approximately Normal? Answer in one sentence.

```{r}
#| label: fig-hist-outcome
#| fig-cap: "Marginal distribution of FDI inflows (% of GDP)."
#| fig-height: 3.5

# Your code here

```

*Your answer:*

## Bivariate scatter plot

Produce a scatter plot of `fdi_pct_gdp` (y-axis) against `gov_effectiveness` (x-axis). Label the axes clearly and give the plot an informative title.

Describe the relationship in **2–3 sentences**: direction, approximate linearity, and the spread of points around the trend.

```{r}
#| label: fig-scatter
#| fig-cap: "Scatter plot of FDI inflows against government effectiveness."
#| fig-height: 4

# Your code here

```

*Your description:*

# Estimand

## Plain language

In **1–2 sentences**, state in plain language what quantity you are trying to estimate in this analysis.

*Your answer:*

## Formal statement

Write the estimand formally. Let $\beta$ denote the population slope, $X_i$ = `gov_effectiveness`, $Y_i$ = `fdi_pct_gdp` for country $i$.

*Your answer:*

## Causal vs. descriptive

Is $\beta$ a causal or a descriptive quantity in this cross-sectional observational dataset? What would be required to interpret it causally? Answer in **2–3 sentences**.

*Your answer:*

# Statistical Model

## Formal model specification

Write out the full statistical model. Your model has three unknown parameters: $\alpha$ (intercept), $\beta$ (slope), and $\sigma$ (error standard deviation).

Use the following priors for this assignment:

$$\alpha \;\sim\; \text{Normal}(3,\; 5) \qquad \beta \;\sim\; \text{Normal}(0,\; 3) \qquad \sigma \;\sim\; \text{Exponential}(0.5)$$

*Write the full likelihood and priors here using the notation from class:*

## Likelihood justification

In **2–3 sentences**, justify the choice of a Gaussian (Normal) likelihood for `fdi_pct_gdp`. What property of the outcome supports this family? How does this connect to Assumption A6 from the Gauss-Markov theorem?

*Your justification:*

## Prior defense

In **2–3 sentences**, explain why these priors are "weakly informative" rather than flat. What does each prior encode about the plausible range of FDI and governance effects? Why is $\beta \sim \text{Normal}(0, 3)$ more scientifically defensible than $\beta \sim \text{Uniform}(-\infty, +\infty)$?

*Your defense:*

# Gauss-Markov Assumptions

## Listing the assumptions

List all six Gauss-Markov assumptions (A1–A6). For each, write:

- one sentence stating what the assumption requires, and
- one sentence describing what a violation would look like in this specific dataset.

*Your answer:*

## Diagnostic plot: linearity and spread

Reproduce your scatter plot from Section 1.4. Does the visual evidence suggest a non-linear relationship (A1) or non-constant spread across the range of `gov_effectiveness` (A3)? Answer briefly.

```{r}
#| label: fig-diagnostic
#| fig-cap: "Scatter plot for Gauss-Markov diagnostic assessment."
#| fig-height: 4

# Copy your scatter code from Section 1 here

```

*Your assessment:*

## The hardest assumption to defend

Which of the six assumptions is most difficult to defend in a cross-sectional observational dataset like this one? In **2–3 sentences**, identify the assumption, explain why it is threatened here, and name at least one specific omitted variable that could pose a problem.

*Your answer:*

# Implementation in brms

## Fitting the model

Fit the Bayesian model from Section 3 using `brm()`. Use 4 chains, 2000 iterations (1000 warmup), and `seed = 2026`. Refer to the `brm()` syntax from the Week 8 slides for the full argument structure. Set `family = gaussian()` explicitly.

```{r}
#| label: fit-model
#| cache: true
#| output: false

# Your code here
# Hint: brm(outcome ~ predictor, data = ..., family = gaussian(),
#           prior = c(...), chains = 4, iter = 2000, warmup = 1000, seed = 2026)

```

## Convergence check

Run `summary()` on your fitted model. Are all $\hat{R}$ values below 1.01? Are all bulk ESS values above 1,000? Briefly comment on what you observe.

```{r}
#| label: model-summary

# summary(your_model_object)

```

*Your convergence assessment:*

## Posterior predictive check

Produce a posterior predictive check using `pp_check()`. In **one sentence**, describe what this plot is showing and whether the model fits the data well.

```{r}
#| label: fig-ppcheck
#| fig-cap: "Posterior predictive check."
#| fig-height: 4

# pp_check(your_model_object, ndraws = 50)

```

*Your interpretation:*

# Model Output

## Posterior summary

Fill in the table below using the values from your `summary()` output.

| Parameter | Posterior Mean | 95% CI Lower | 95% CI Upper |
|---|---|---|---|
| $\alpha$ (Intercept) | | | |
| $\beta$ (`gov_effectiveness`) | | | |

## Interpret the intercept

Interpret $\alpha$ in plain language. What does the model's intercept represent in the context of this study?

*Your interpretation:*

## Interpret the slope

Interpret $\beta$ in plain language. How many additional percentage points of FDI inflows are associated with a one-unit increase in `gov_effectiveness`? Is this effect substantively meaningful given the scale of the outcome?

*Your interpretation:*

## Credible interval interpretation

Write **one sentence** interpreting the 95% credible interval for $\beta$ as a direct probability statement.

*Your answer:*

# Visualization and Substantive Interpretation

## Cloud of posterior regression lines

Extract 200 samples from the posterior using `as_draws_df()`. Plot these as semi-transparent regression lines over the scatter plot of the data. Add the posterior mean regression line in a distinct colour.

::: {.callout-note}
After `as_draws_df(your_model_object)`, the posterior intercept column is named `b_Intercept` and the slope column is `b_gov_effectiveness`.
:::

```{r}
#| label: fig-posterior-lines
#| fig-cap: "200 posterior draws as plausible regression lines over the data. Each grey line is one plausible regression consistent with the data and priors."
#| fig-height: 4

# Hint: use geom_abline() with data = post_draws for the grey lines,
#       then geom_abline() again for the posterior mean, then geom_point() for the data.

```

## Marginal posterior of the slope

Plot the marginal posterior distribution of `b_gov_effectiveness` as a density curve. Mark the posterior mean with a solid vertical line, the 95% CI boundaries with dashed vertical lines, and zero with a dotted vertical line for reference.

```{r}
#| label: fig-slope-posterior
#| fig-cap: "Marginal posterior distribution of the slope coefficient."
#| fig-height: 3.5

# Your code here

```

## Substantive paragraph

In **2–3 sentences**, write a substantive interpretation of your results for a policy analyst who has never seen a posterior distribution. What does the analysis tell us about the relationship between governance and FDI? Include **one important caveat**.

*Your paragraph:*

---

# Submission Checklist

Before submitting, verify that your rendered PDF:

- [ ] Contains your name in the YAML header above
- [ ] Includes descriptive statistics and both exploratory plots (Section 1)
- [ ] States the estimand formally and discusses its causal vs. descriptive nature (Section 2)
- [ ] Writes the full likelihood and three prior distributions (Section 3)
- [ ] Lists all six Gauss-Markov assumptions with explanations and a diagnostic scatter plot (Section 4)
- [ ] Fits the `brms` model and reports convergence diagnostics and a `pp_check()` (Section 5)
- [ ] Reports and interprets posterior mean and 95% CI for $\alpha$ and $\beta$ (Section 6)
- [ ] Includes the cloud-of-lines plot, slope posterior plot, and substantive paragraph (Section 7)
- [ ] Was produced by rendering this `.qmd` file without errors
