---
title: "Homework 05: Visualising democracy with V-Dem"
author: "Your name"
date: last-modified
date-format: MMMM D, YYYY
format:
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup

start_time <- Sys.time()

# Install once, in the Console, not here:
# install.packages("pak")
# pak::pak("vdeminstitute/vdemdata")

library(tidyverse) # wrangling and visualization
library(janitor) # quick tables with tabyl()
library(vdemdata) # the V-Dem dataset

set.seed(2026)

# One row per country and year, from 1950 on.
vdem_df <- vdem |>
  as_tibble() |>
  filter(year >= 1950) |>
  transmute(
    country = country_name,
    year,
    continent = case_match(
      e_regiongeo,
      1:4 ~ "Europe",
      5:9 ~ "Africa",
      10:14 ~ "Asia",
      15 ~ "Oceania",
      16:19 ~ "Americas"
    ),
    regime = factor(
      case_match(
        v2x_regime,
        0 ~ "Closed autocracy",
        1 ~ "Electoral autocracy",
        2 ~ "Electoral democracy",
        3 ~ "Liberal democracy"
      ),
      levels = c(
        "Closed autocracy",
        "Electoral autocracy",
        "Electoral democracy",
        "Liberal democracy"
      )
    ),
    libdem = v2x_libdem,
    polyarchy = v2x_polyarchy,
    liberal = v2x_liberal,
    egaldem = v2x_egaldem,
    partipdem = v2x_partipdem,
    freexp = v2x_freexp_altinf,
    corruption = v2x_corr,
    women = v2x_gender,
    gdp_pc = e_gdppc,
    pop = e_pop / 100
  )
```

# The data {#sec-data}

This homework repeats the week 5 session on new data.
Instead of gapminder, you explore how democracy has changed since 1950, over time and across continents.

The data come from **V-Dem** (Varieties of Democracy), which measures democracy in almost every country of the world since 1789.
The package `vdemdata` ships the whole dataset as one data frame, `vdem`, with more than 4,000 columns.
The setup chunk above has already selected the columns you need, renamed them and stored them in `vdem_df`.
Read it before you start.
`transmute()` works like `mutate()` but keeps only the columns it creates, and `case_match()` turns V-Dem's number codes into readable categories.

| Column | What it measures | Scale |
|---|---|---|
| `country`, `year` | The country and the year | |
| `continent` | Built from V-Dem's geographic region, like gapminder's `continent` | Five categories |
| `regime` | Regimes of the World: closed autocracy, electoral autocracy, electoral democracy, liberal democracy | Ordered categories |
| `libdem` | Liberal democracy index: free elections plus civil liberties, rule of law and checks on the executive | 0 (low) to 1 (high) |
| `polyarchy` | Electoral democracy index: free and fair elections | 0 to 1 |
| `liberal` | Liberal component: civil liberties, rule of law, checks on the executive | 0 to 1 |
| `egaldem` | Egalitarian democracy index | 0 to 1 |
| `partipdem` | Participatory democracy index | 0 to 1 |
| `freexp` | Freedom of expression and alternative sources of information | 0 to 1 |
| `corruption` | Political corruption index, where **higher means more corrupt**, unlike every other index | 0 to 1 |
| `women` | Women's political empowerment index | 0 to 1 |
| `gdp_pc` | GDP per capita, a model-based estimate whose unit is not US dollars, so compare countries by ratios rather than differences | Ends in 2019 |
| `pop` | Population in millions, a model-based estimate | Ends in 2019 |

: The columns of `vdem_df`, built in the setup chunk from the V-Dem dataset. {#tbl-vdem-columns}

Use AI tools to explain an error message or a function, not to write your answers.
The point of this homework is the practice.

# Explore {#sec-explore}

1. `glimpse()` `vdem_df`.
   How many rows and columns does it have, and what is one row?
2. How many countries does it cover, and which years?
   `count(country)` returns one row per country, as in homework 04, so `nrow()` of it is the number of countries.
   `summarise()` with `min()` and `max()` returns the first and the last year.

```{r}
#| label: ex-explore
```

3. Answer in prose: data that observe the same units again and again over time are called **panel data**.
   Name one question you could answer with `vdem_df` that you could not answer with data from a single year.

4. Create `vdem_2025`, the rows of the most recent year, and make a quick table of `regime` with `tabyl()` and `adorn_totals()`, as in class.
   Which regime type is the most common in 2025?
5. `gdp_pc` and `pop` end in 2019.
   Check this: count the non-missing values of `gdp_pc` in every year from 2016 to 2022.
   Hint: `summarise(n_gdp = sum(!is.na(gdp_pc)))` after a `group_by(year)`.
6. Create `vdem_2019`, the rows of 2019.
   Every figure that uses `gdp_pc` or `pop` uses `vdem_2019`.

```{r}
#| label: ex-explore-years
```

# The grammar of graphics {#sec-grammar}

1. Build a scatter plot of `libdem` against `gdp_pc` in `vdem_2019` the way we did in class, one layer at a time: the data, then the aesthetics, then a geom, then `scale_x_log10()`.
   Run the code after every layer, and keep only the final figure in the chunk.

```{r}
#| label: ex-grammar-build
```

2. In one sentence each: what did each of the four layers add?
3. Why does a log scale suit GDP per capita?
   Look at the figure with and without it.

4. Draw the same scatter plot twice: once with every point in one fixed colour of your choice, and once with each continent in its own colour.
   Which of the two sets `colour` inside `aes()`, and which outside?

```{r}
#| label: ex-grammar-colour-outside
```

```{r}
#| label: ex-grammar-colour-inside
```

5. **Before running anything**, predict what `aes(colour = "blue")` does.
   Then run it and explain what you see.

```{r}
#| label: ex-grammar-colour-blue
```

# One variable {#sec-one-variable}

1. Draw a bar chart of `regime` in `vdem_2025`.
   The bars follow the order of the factor levels set in the setup chunk, from the least to the most democratic regime.
   Here the alphabet happens to give the same order.
   Why is it still worth setting the order explicitly?

```{r}
#| label: ex-bar
```

2. The category names overlap on the x-axis.
   Map `regime` to `y` instead of `x` and see what happens.

```{r}
#| label: ex-bar-horizontal
```

3. Draw a histogram of `libdem` in `vdem_2025`, once with `binwidth = 0.1` and once with `binwidth = 0.02`.
   Describe the shape of the distribution in one or two sentences.
   Which bin width shows it better, and why?

```{r}
#| label: ex-histogram-wide
```

```{r}
#| label: ex-histogram-narrow
```

4. Draw a density plot of `libdem` in `vdem_2025`, filled with a colour of your choice, with `alpha = 0.5`.
   What does the density plot show less clearly than the histogram?

```{r}
#| label: ex-density
```

5. **Stretch:** compare the distribution of `libdem` in 1975 with the one in 2025 in one density plot.
   Hint: filter both years, and map `fill = factor(year)`.

```{r}
#| label: ex-density-stretch
```

# Two variables {#sec-two-variables}

1. Draw a scatter plot of `libdem` against `corruption` in `vdem_2025`.
   Describe the relationship in one or two sentences, and remember which way `corruption` runs.

```{r}
#| label: ex-scatter
```

2. Draw a box plot of `libdem` by `continent` in `vdem_2025`.
   Then draw the same data as a violin, with the countries as jittered points on top.
   What does the violin with points show that the box plot hides?

```{r}
#| label: ex-boxplot
```

```{r}
#| label: ex-violin
```

3. Pick three countries that interest you and draw a line plot of their `libdem` over time, one colour per country.
   Hint: `%in%` and `colour = country`, as in class.
   Describe one turning point you see, and what you think happened.

```{r}
#| label: ex-line
```

# Three and more variables {#sec-more-variables}

1. Extend the scatter plot of `libdem` against `gdp_pc` in `vdem_2019`: map `continent` to colour and `pop` to point size.
   How many variables does the figure show now?

```{r}
#| label: ex-colour-size
```

2. Encode `continent` redundantly, with colour and point shape.
   Why does that help?

```{r}
#| label: ex-redundant
```

3. Plot `libdem` over time for **every** country, one line per country, with one panel per continent.
   Use `group = country` and a low `alpha`.
4. **Stretch:** draw one country of your choice on top, in a colour that stands out against the others.
   Hint: a second `geom_line()` with its own `data =`.

```{r}
#| label: ex-facets
```

5. In two sentences: which continents saw the largest changes, and when?

# Styling: less ink, more information {#sec-styling}

1. Make the faceted line plot from @sec-more-variables publication-ready:
   - a title that states a finding, and a subtitle that says what is plotted
   - axis labels with units, or no label where the axis speaks for itself
   - a caption with the source: `"Source: V-Dem dataset v16."`
   - `theme_classic()`
   - if you did the stretch, a label for your country with `geom_text()`

```{r}
#| label: ex-styled-facets
```

2. Annotate the scatter plot of `libdem` against `gdp_pc` in `vdem_2019`:
   - every country in one muted colour, such as a grey
   - a dashed horizontal line at the median of `libdem`, labelled with `annotate()`
   - one country of your choice highlighted in a contrasting colour, labelled with its name
   - `theme_classic()` and readable labels

```{r}
#| label: ex-annotate
```

3. Redo the scatter plot with colour and size from @sec-more-variables with a colourblind-safe palette: `scale_colour_viridis_d(end = 0.85)`.
   Why `end = 0.85`?
   Try the plot without it.

```{r}
#| label: ex-viridis
```

# Common pitfalls: what is wrong here? {#sec-pitfalls}

The two figures below are deliberately bad.
Run them, and use the checklist from the end of the week 5 slides.

```{r}
#| label: fig-bad-lines
#| fig-cap: "A deliberately bad figure: mean liberal democracy by continent since 1950."

vdem_df |>
  group_by(continent, year) |>
  summarise(mean_libdem = mean(libdem, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = year, y = mean_libdem, colour = continent)) +
  geom_line(linewidth = 1.5) +
  scale_colour_manual(
    values = c("red3", "green4", "darkorange3", "forestgreen", "firebrick")
  ) +
  theme_dark()
```

```{r}
#| label: fig-bad-bars
#| fig-cap: "A deliberately bad figure: mean liberal democracy by continent in 2025."

vdem_df |>
  filter(year == 2025) |>
  group_by(continent) |>
  summarise(mean_libdem = mean(libdem, na.rm = TRUE)) |>
  ggplot(aes(x = continent, y = mean_libdem, fill = continent)) +
  geom_col() +
  coord_cartesian(ylim = c(0.2, 0.6))
```

1. List at least two problems in @fig-bad-lines and at least two in @fig-bad-bars.
   For each, say what a reader would get wrong.
2. Fix both figures.
   The cells below already carry figure labels.
   Replace each placeholder `fig-cap` with a caption that a reader can understand without the text.

```{r}
#| label: fig-fix-lines
#| fig-cap: "Replace this with your caption."
```

```{r}
#| label: fig-fix-bars
#| fig-cap: "Replace this with your caption."
```

# Your figure: a data summary {#sec-figure}

::: {.callout-important}
**This section is step 2 of the workflow, the data summary.**
In week 4 the data summary was a table.
Here it is a figure that answers a precise question, and a reader should understand it without your code.
:::

**The question.**
How has the share of countries in each regime type changed since 1950, and does the pattern differ by continent?

1. Write **one pipeline**, starting from `vdem_df`, that:
   - drops the rows without a `regime`
   - counts the countries per `continent`, `year` and `regime`
   - turns each count into a share of the countries on that continent in that year

   Assign the result to `regime_shares`, and run the pipeline after every new line, as in class.
   Hint: `count(continent, year, regime)`, then `group_by(continent, year)` and `mutate(share = n / sum(n))`.
   End with `complete(continent, year, regime, fill = list(n = 0, share = 0))`: a regime type that no country had in some year has no row, and `geom_area()` needs one with a share of zero.

```{r}
#| label: ex-regime-shares
```

2. Plot `regime_shares` as a figure for a reader: the share on the y-axis, the year on the x-axis, `regime` as the fill colour, and one panel per continent.
   `geom_area()` stacks the shares on top of each other, as a stacked bar chart stacks counts.
   Style it as in @sec-styling, with a viridis palette.
   The cell below already carries the label `fig-regimes`.
   Replace its placeholder `fig-cap` with a caption that a reader can understand without the text.

```{r}
#| label: fig-regimes
#| fig-cap: "Replace this with your caption."
```

3. Below the figure, write three to four sentences that answer the question.
   - Reference the figure with `@fig-regimes`.
   - Put at least one number into a sentence with inline code, so that it updates when the data does.
   - Name one thing the figure hides, or one reason to read it with care.
     Think about what one country counts for (India and Fiji count the same), and about which units V-Dem counts in 1950, when much of Africa and Asia was still under colonial rule.

# Reflection {#sec-reflection .unnumbered}

Which figure took you longest, and why?
Is there a part of `ggplot2` you are still unsure about?

# Session info and execution time {#sec-session .unnumbered}

```{r}
#| label: session-info

sessionInfo()
```

```{r}
#| label: exec-time

end_time <- Sys.time()
cat(
  "Total execution time:",
  round(difftime(end_time, start_time, units = "mins"), 1),
  "minutes\n"
)
```
