---
title: "Week 4: The tidyverse"
---

<!-- Shared format options live in _metadata.yml, so this deck's YAML carries only the title.

     This week teaches code, so the code has to be visible: every teaching chunk opts out of the
     default fold with #| code-fold: false, as in week 03.

     Rhythm: every concept is followed by a "Practice:" slide the room does on the spot, with the
     answer behind a . . . beat. The red line is ONE story: read a messy file, make it tidy, then
     transform, join and summarise, while a single gapminder pipeline grows one verb at a time
     (the new line is highlighted with code-line-numbers). It ends in a pipeline the room writes.

     Time budget (90 min): recap 8, errors 5, packages 10, tidyverse and tidy data 10, read 8,
     tidy 8, transform 17, join 7, summarise 9, final challenge 6, wrap-up 2.
     If the room runs slow, cut arrange first, then pivot_wider.

     data/week04-lifeexp-wide.csv is gapminder's life expectancy for the nine countries of the
     hw-03 `europe` table, pivoted wide (one column per year) and written once with:
       gapminder |> filter(country %in% nine) |> select(country, year, lifeExp) |>
         pivot_wider(names_from = year, values_from = lifeExp) |> write_csv(...)
     Students read the same file from its raw GitHub URL, so it must be pushed before class. -->

```{r}
#| label: setup
#| include: false

start_time <- Sys.time()

library(tidyverse) # wrangling
library(here) # project-root-relative paths
library(gapminder) # demo data

# _quarto.yml is a project-root marker, so anchor here() at the repo root.
here::i_am("website/slides/week04.qmd")

# Short tibble printouts, so a result fits under its code on a slide.
options(pillar.print_min = 5, pillar.print_max = 5, width = 70)
```

## Week 4 goals {#sec-week04-goals .smaller}

By the end of today, you will be able to:

1. Use Quarto cell options to control what a rendered document shows.
2. Tell an R error, a Quarto error and a LaTeX error apart.
3. Install and load packages, and call a function as `package::function()`.
4. Say what the tidyverse is and what makes data tidy.
5. Read a file and reshape it into tidy form.
6. Transform, join and summarise data in one readable pipeline.

## Today {#sec-today .smaller}

1. **Recap:** Quarto cell options
2. **Errors:** which layer is complaining?
3. **Packages:** install, load, `package::function()`
4. **The tidyverse** and tidy data
5. **Read, tidy, transform:** one dataset, one growing pipeline
   - `read_csv()`, `pivot_longer()`, the `dplyr` verbs, `left_join()`, `summarise()`
6. **Your pipeline**

::: notes
Same deal as last week: type along.
Positron open, a fresh `.qmd` or the console.
:::

## Where today fits {#sec-where-today-fits .smaller}

Today starts **step 2, the data summary**.
Every data summary you will write this term is a pipeline like the one we build today.

{{< include _workflow-8step.qmd >}}

::: notes
One minute.
Point at step 2 only.
:::

# Recap: Quarto cell options {#sec-cell-options}

## Cell options you have met {#sec-cell-options-table .smaller}

::: notes
Two minutes.
They used label, echo, eval and error in hw-03.
Add include, message, warning and fig-cap as the new ones.
:::

:::: {.columns}

::: {.column width="60%"}
::: {.small}
| Option | What it does |
|---|---|
| `label: my-name` | Names the cell. Unique, and required. |
| `echo: false` | Runs the code, hides it, shows the result. |
| `eval: false` | Shows the code, does not run it. |
| `include: false` | Runs the code, shows nothing at all. |
| `error: true` | Shows an error and keeps rendering. |
| `message: false` | Hides messages, such as package startup text. |
| `warning: false` | Hides warnings. |
| `fig-cap: "..."` | Captions a figure, with a `fig-` label. |

: The cell options this course uses, one `#|` line each at the top of a cell. {#tbl-cell-options}
:::
:::

::: {.column width="40%"}
Options set in the YAML header apply to **every** cell.
A cell option overrides them for that one cell.

```yaml
execute:
  echo: true
  message: false
  warning: false
```
:::

::::

## Practice: predict the PDF {#sec-practice-cell-options .smaller}

**Do this now.**
For each cell, what appears in the rendered document: code, result, both, or nothing?

:::: {.columns}

::: {.column width="50%"}
**A**

```{{r}}
#| label: a
#| echo: false

mean(c(2, 4, 6))
```

**B**

```{{r}}
#| label: b
#| eval: false

mean(c(2, 4, 6))
```
:::

::: {.column width="50%"}
**C**

```{{r}}
#| label: c
#| include: false

avg <- mean(c(2, 4, 6))
```

**D**

```{{r}}
#| label: d
#| error: true

mean(c(2, 4, 6)) + "one"
```
:::

::::

. . .

A: only the result `4`.
B: only the code.
C: nothing, but `avg` now exists for later cells.
D: the code and the error, and the render goes on.

# Three layers of errors {#sec-error-layers}

## Rendering is a relay {#sec-render-relay .smaller}

::: notes
Three minutes.
The one idea: the first word of the message tells you which runner dropped the baton.
Week 3 taught reading an R error, this is the step before: is it R at all?
:::

`.qmd` → **R** runs the cells → `.md` → **Quarto and Pandoc** convert → `.tex` → **LaTeX** typesets → `.pdf`

. . .

Each runner has its own voice when it fails:

::: {.small}
| Layer | The message looks like | Where to look |
|---|---|---|
| **R** | `Error:` or `Error in f():`, then `Quitting from hw.qmd:6-10 [cell-label]` | The cell named in square brackets, at those lines. |
| **Quarto** | `ERROR: YAMLException: ... (6:20)` | The YAML header, at line 6, column 20. |
| **LaTeX** | `compilation failed- error`, then a line number `l.172` | Your prose: a special character such as `%`, `&`, `$` or a stray `\`. |

: The three layers of a PDF render, how each one reports a failure, and where to go to fix it. {#tbl-error-layers}
:::

::: {.callout-tip}
**Only the first error counts.**
Scroll up to the first message, fix that, and render again.
:::

## Practice: which layer? {#sec-practice-error-layers .smaller}

**Do this now.**
Real messages from real renders.
Which layer broke, and what would you check first?

:::: {.columns}

::: {.column width="50%"}
**1**

```text
compilation failed- error
Undefined control sequence.
l.172 GDP grew by 5\% and the \gdp
```

**2**

```text
ERROR: YAMLException: bad indentation
of a mapping entry (6:20)
 5 |    toc: true
 6 |     number-sections: true
```
:::

::: {.column width="50%"}
**3**

```text
Error:
! object 'lifeexp' not found
Quitting from hw-04.qmd:6-10 [life-mean]
```
:::

::::

. . .

1: **LaTeX**, the prose holds `\gdp`, which LaTeX reads as a command.
2: **Quarto**, the YAML header is misindented at line 6.
3: **R**, in the cell `life-mean`: a typo or a line not run yet, as in week 3.

# Packages {#sec-packages}

## Packages: functions other people wrote {#sec-what-packages .smaller}

::: notes
Callback to week 3: they wrote `celsius_to_fahrenheit()`.
A package is that, times a few hundred, with documentation, shared.
:::

Last week you wrote your own function.
A **package** is a bundle of functions, their help pages, and sometimes data, that someone wrote and shared.

. . .

- R itself ships with the **base** packages: `mean()`, `round()`, `data.frame()`.
- **CRAN**, R's package archive, holds more than 20,000 more, checked and free.
- Nearly every method in this course lives in a package: `dplyr` today, `ggplot2` next week, `brms` later.

## Install once, load every session {#sec-install-load .smaller}

:::: {.columns}

::: {.column width="50%"}
**Install** downloads the package to your computer.
Once, in the **console**.

```{r}
#| label: install-demo
#| code-fold: false
#| eval: false

install.packages("gapminder")
```
:::

::: {.column width="50%"}
**Load** makes its functions available.
Every session, at the top of your `.qmd`.

```{r}
#| label: load-demo
#| code-fold: false

library(gapminder) # demo data
```
:::

::::

. . .

Like an app on your phone: you install it once, and you open it every time you use it.

::: {.callout-important}
**`install.packages()` never goes into a `.qmd`.**
It would reinstall on every render, and Quarto refuses to run it anyway.
:::

## Quarto stops when a package is missing {#sec-missing-package .smaller}

If a `library()` call names a package that is not installed, the render stops at that cell:

```text
Error in `library()`:
! there is no package called 'gapminder'
Quitting from hw-04.qmd:6-10 [setup]
```

. . .

- That is an **R** error, reported through Quarto.
- The fix is in the **console**, not in the document: `install.packages("gapminder")`, then render again.
- It also happens when the name is misspelled: `library(gapmindr)` gives the same message.

::: notes
This is the most common first-render failure on a new laptop.
Everybody who installed a package in a different R version meets it once.
:::

## `package::function()` {#sec-double-colon .smaller}

`::` names the package a function comes from.

```{r}
#| label: double-colon
#| code-fold: false

gapminder::gapminder |> nrow()
```

. . .

Three reasons to use it:

- **Clarity:** the reader sees where a function lives.
- **No `library()` needed** for a one-off call, such as `readxl::read_excel()`.
- **Name clashes:** two packages can both have a `filter()`.
  `library(tidyverse)` tells you which one wins:

```text
✖ dplyr::filter() masks stats::filter()
```

## Practice: install and load {#sec-practice-packages .smaller}

**Do this now.**

1. In the **console**, install `gapminder`.
2. At the top of your `.qmd`, load `tidyverse` and `gapminder` in a setup cell.
   Run the cell and read the startup message.
   Which packages were attached, and what is masked?
3. Type `gapminder` and run it.
   How many rows and columns?
4. Run `?gapminder`.
   What does `gdpPercap` measure?

. . .

```{r}
#| label: practice-packages
#| code-fold: false

library(tidyverse) # wrangling
library(gapminder) # demo data

gapminder
```

::: notes
Circulate.
Anyone whose `library(tidyverse)` fails installs it now.
:::

# The tidyverse {#sec-tidyverse}

## One grammar across many packages {#sec-tidyverse-philosophy .smaller}

The **tidyverse** is a family of packages built on one shared philosophy:

. . .

- **Tidy data in, tidy data out:** every function takes a table and returns a table.
- **Data first:** the table is always the first argument, so every step chains with `|>`.
- **Functions are verbs:** `filter()`, `select()` and `summarise()` read like what they do.
- **Written for humans:** consistent names and arguments, so learning one package teaches you the next.

. . .

`library(tidyverse)` loads the core packages in one line.

::: notes
Two minutes.
The payoff of the design: tomorrow's package already feels familiar.
:::

## What is tidy data? {#sec-tidy-data .smaller}

> Tidy datasets are all alike, but every messy dataset is messy in its own way.
>
> Hadley Wickham

. . .

A dataset is **tidy** when:

1. Each **variable** has its own **column**.
2. Each **observation** has its own **row**.
3. Each **value** has its own **cell**.

. . .

This is the data frame from week 3, with one rule added: **decide what a row is**.
In gapminder, a row is one country in one year.

## Practice: is this tidy? {#sec-practice-tidy .smaller}

**Do this now.**
Tidy or not?
If not, which rule does it break?

:::: {.columns}

::: {.column width="50%"}
::: {.small}
| country | year | lifeExp |
|---|---|---|
| Germany | 1952 | 67.5 |
| Germany | 2007 | 79.4 |

: Layout A. {#tbl-layout-a}

| country | 1952 | 2007 |
|---|---|---|
| Germany | 67.5 | 79.4 |
| France | 67.4 | 80.7 |

: Layout B. {#tbl-layout-b}
:::
:::

::: {.column width="50%"}
::: {.small}
| country | year | measure | value |
|---|---|---|---|
| Germany | 2007 | lifeExp | 79.4 |
| Germany | 2007 | pop | 82400996 |

: Layout C. {#tbl-layout-c}

| country | year | life_range |
|---|---|---|
| Germany | 1952-2007 | 67.5-79.4 |

: Layout D. {#tbl-layout-d}
:::
:::

::::

. . .

A is tidy.
B hides the variable `year` in its column names.
C stores two variables in rows.
D puts two values in each cell.

::: notes
Hands up for each one before revealing.
B is the one we meet on the next slides: it is our file.
:::

## The core packages, along the data cycle {#sec-core-packages .smaller}

**Read** → **Tidy** → **Transform** → Visualise → Model → Communicate

::: {.small}
| Stage | Package | For | Key functions |
|---|---|---|---|
| Read | `readr` | Text files such as CSV | `read_csv()`, `write_csv()` |
| Tidy | `tidyr` | Reshaping | `pivot_longer()`, `pivot_wider()` |
| Transform | `dplyr` | The verbs of data manipulation | `filter()`, `mutate()`, `summarise()` |
| Transform | `stringr`, `forcats`, `lubridate` | Text, factors, dates | `str_detect()`, `fct_reorder()`, `ymd()` |
| Visualise | `ggplot2` | Plots, next week | `ggplot()` |
| Underneath | `tibble`, `purrr` | Data frames, repetition | `tibble()`, `map()` |

: The core tidyverse packages, ordered along the data cycle. Today covers the first three stages. {#tbl-core-packages}
:::

::: notes
Do not read the table.
Point at the three stages we cover, then move on.
:::

# Read {#sec-read}

## Every format has a reader {#sec-import-formats .smaller}

::: {.small}
| Format | Extension | Function | Loaded by `library(tidyverse)`? |
|---|---|---|---|
| Comma-separated | `.csv` | `readr::read_csv()` | yes |
| Semicolon-separated, as German Excel saves it | `.csv` | `readr::read_csv2()` | yes |
| Tab-separated | `.tsv` | `readr::read_tsv()` | yes |
| R's own format | `.rds` | `readr::read_rds()` | yes |
| Excel | `.xlsx`, `.xls` | `readxl::read_excel()` | installed, not loaded |
| Stata | `.dta` | `haven::read_dta()` | installed, not loaded |
| SPSS | `.sav` | `haven::read_sav()` | installed, not loaded |
| JSON | `.json` | `jsonlite::fromJSON()` | installed, not loaded |
| Parquet | `.parquet` | `arrow::read_parquet()` | install it yourself |

: Common data formats and the function that reads each one. Installing the tidyverse also installs `readxl`, `haven` and `jsonlite`, but only `library(tidyverse)` loads `readr`, which is where `::` earns its keep. {#tbl-import-formats}
:::

::: notes
Point out the German Excel row: a CSV that comes out as one column is usually a semicolon file.
The next homework reads an `.xlsx` with `readxl::read_excel()`.
:::

## Reading a file {#sec-read-csv .smaller}

```{r}
#| label: read-csv
#| code-fold: false

lifeexp_wide <- read_csv(here("website", "slides", "data", "week04-lifeexp-wide.csv"))
glimpse(lifeexp_wide)
```

. . .

- **Assign the result.** `read_csv()` alone prints the data and forgets it.
- `here()` builds the path from your project folder, so it works on any computer.
  In your homework repository that is `here("data", "parlgov.xlsx")`.
- `glimpse()` shows every column with its type: the first thing to run on new data.

## Practice: read it yourself {#sec-practice-read .smaller}

**Do this now.**
`read_csv()` also reads straight from the web.

```{r}
#| label: practice-read
#| code-fold: false
#| eval: false

url <- "https://raw.githubusercontent.com/qmir-2026-fall/qmir-2026-fall.github.io/main/website/slides/data/week04-lifeexp-wide.csv"
lifeexp_wide <- read_csv(url)
```

1. Read the file and `glimpse()` it.
2. Which rule of tidy data does it break?
3. A tidy version has one row per country and year.
   How many rows will it have?

. . .

It is Layout B: `year` hides in the column names.
Nine countries times twelve years makes **108 rows**.

# Tidy {#sec-tidy}

## `pivot_longer()`: wide to tidy {#sec-pivot-longer .smaller}

```{r}
#| label: pivot-longer
#| code-fold: false

lifeexp <- lifeexp_wide |>
  pivot_longer(
    cols = -country, # every column except country
    names_to = "year", # the column names become this variable
    values_to = "lifeExp", # the cell values become this one
    names_transform = as.integer # "1952" is text, make it a number
  )
lifeexp
```

::: notes
Read the call aloud as a sentence: take every column except country, put the names into year and the values into lifeExp.
:::

## `pivot_wider()`: the way back {#sec-pivot-wider .smaller}

Wide is not wrong.
It is how a **table for a reader** usually looks, and `pivot_wider()` gets you there:

```{r}
#| label: pivot-wider
#| code-fold: false

lifeexp |>
  pivot_wider(names_from = year, values_from = lifeExp)
```

. . .

::: {.callout-tip}
**Tidy for the computer, wide for the reader.**
Keep your working data tidy, and pivot wider only at the very end, for a table in your report.
:::

## Practice: pivot it {#sec-practice-pivot .smaller}

**Do this now.**

1. Pivot `lifeexp_wide` into a tidy tibble called `lifeexp`.
2. Check with `nrow()` that it has 108 rows.
3. Check with `class(lifeexp$year)` that `year` is a number.
   What happens without `names_transform`?
4. **Stretch:** pivot it back to wide.

. . .

Without `names_transform`, `year` stays **character**, and `year > 2000` would compare text.
From here on we use the full `gapminder`: it is this tibble for all 142 countries, with three more variables.

::: notes
The transition to gapminder is the key sentence here.
Our file was a slice of gapminder all along.
:::

# Transform {#sec-transform}

## The `dplyr` verbs {#sec-dplyr-verbs .smaller}

:::: {.columns}

::: {.column width="55%"}
::: {.small}
| Verb | What it does |
|---|---|
| `filter()` | Keep rows that meet a condition |
| `select()` | Keep or drop columns |
| `mutate()` | Create or change columns |
| `arrange()` | Sort rows |
| `left_join()` | Add columns from another table |
| `group_by()` | Split rows into groups |
| `summarise()` | Collapse each group to one row |

: The core `dplyr` verbs covered today, in the order the pipeline uses them. {#tbl-dplyr-verbs}
:::
:::

::: {.column width="45%"}
Every verb:

- takes a **table first**, so it pipes
- names columns **without quotes** and without `$`
- returns a **new table** and leaves the old one unchanged
:::

::::

. . .

We build **one pipeline**, one verb at a time.
The highlighted line is the new one.

## `filter()`: keep rows {#sec-filter .smaller}

```{r}
#| label: pipe-filter
#| code-fold: false
#| code-line-numbers: "2"

gapminder |>
  filter(continent == "Europe", year == 2007)
```

. . .

- A comma means **and**.
  `|` means or, and `%in%` tests membership, as in week 3.
- `==` compares, `=` names an argument.

## `select()`: keep columns {#sec-select .smaller}

```{r}
#| label: pipe-select
#| code-fold: false
#| code-line-numbers: "3"

gapminder |>
  filter(continent == "Europe", year == 2007) |>
  select(country, lifeExp, pop, gdpPercap)
```

. . .

A minus drops a column instead: `select(-continent)`.

## Practice: filter and select {#sec-practice-filter-select .smaller}

**Do this now.**

1. Keep the Asian countries in 1952, with only `country` and `lifeExp`.
2. Which countries had a life expectancy above 80 in 2007?
3. **Stretch:** keep only the nine countries of our file, in every year.
   Hint: `%in%` and the `country` column of `lifeexp_wide`.

. . .

```{r}
#| label: practice-filter-select
#| code-fold: false

gapminder |>
  filter(lifeExp > 80, year == 2007) |>
  select(country, lifeExp)
```

`filter(country %in% lifeexp_wide$country)` answers the stretch.

## `mutate()`: new columns {#sec-mutate .smaller}

```{r}
#| label: pipe-mutate
#| code-fold: false
#| code-line-numbers: "4-7"

gapminder |>
  filter(continent == "Europe", year == 2007) |>
  select(country, lifeExp, pop, gdpPercap) |>
  mutate(
    pop_m = pop / 1e6,
    gdp_bn = pop * gdpPercap / 1e9
  )
```

::: notes
`1e6` is a million, `1e9` a billion.
Several new columns in one `mutate()`, separated by commas.
:::

## `arrange()`: sort rows {#sec-arrange .smaller}

```{r}
#| label: pipe-arrange
#| code-fold: false
#| code-line-numbers: "8"

gapminder |>
  filter(continent == "Europe", year == 2007) |>
  select(country, lifeExp, pop, gdpPercap) |>
  mutate(
    pop_m = pop / 1e6,
    gdp_bn = pop * gdpPercap / 1e9
  ) |>
  arrange(desc(gdp_bn))
```

. . .

`arrange()` sorts ascending, and `desc()` flips it.

## Practice: mutate and arrange {#sec-practice-mutate-arrange .smaller}

**Do this now.**

1. For 2007, add a column `pop_m` with the population in millions.
2. Sort so that the most populous country comes first.
   Which five countries lead?
3. **Stretch:** which country gained the most life expectancy between 1952 and 2007?
   Use your tidy `lifeexp`, `pivot_wider()` and `mutate()`.

. . .

```{r}
#| label: practice-mutate-arrange
#| code-fold: false

gapminder |>
  filter(year == 2007) |>
  mutate(pop_m = pop / 1e6) |>
  arrange(desc(pop_m)) |>
  select(country, pop_m)
```

::: notes
Stretch answer: pivot_wider, then mutate(gain = `2007` - `1952`), then arrange(desc(gain)).
Turkey, by about 28 years.
:::

# Join {#sec-join}

## `left_join()`: add columns from another table {#sec-left-join .smaller}

gapminder does not know who is in the EU.
The `europe` table from hw-03 does:

:::: {.columns}

::: {.column width="45%"}
```{r}
#| label: eu-accession
#| code-fold: false

eu_accession <- tibble(
  country = c(
    "France", "Germany", "Italy",
    "Spain", "Sweden", "Poland"
  ),
  eu_joined = c(
    1958, 1958, 1958,
    1986, 1995, 2004
  )
)
```
:::

::: {.column width="55%"}
```{r}
#| label: pipe-join
#| code-fold: false
#| code-line-numbers: "4"

gapminder |>
  filter(country %in% lifeexp_wide$country) |>
  filter(year == 2007) |>
  left_join(eu_accession, by = "country") |>
  select(country, lifeExp, eu_joined)
```
:::

::::

::: notes
Before running the join, ask the prediction question on the next slide.
:::

## Practice: predict the join {#sec-practice-join .smaller}

**Do this now.**
Before you run the join, write down:

1. How many rows will the result have?
2. What will `eu_joined` hold for Norway?
3. What happens to a country spelled `"Czechia"` in one table and `"Czech Republic"` in the other?

. . .

1. **Nine**: `left_join()` keeps every row of the left table.
2. **`NA`**: Norway has no match, and `NA` is R's missing value from week 3.
3. It silently gets `NA` too.

::: {.callout-warning}
**A join matches text exactly.**
One spelling difference and the row gets `NA`, without any error.
Always check how many `NA`s a join produced.
:::

# Summarise {#sec-summarise}

## `summarise()`: many rows to one {#sec-summarise-intro .smaller}

:::: {.columns}

::: {.column width="50%"}
```{r}
#| label: summarise-all
#| code-fold: false

gapminder |>
  filter(year == 2007) |>
  summarise(
    mean_life = mean(lifeExp),
    n = n()
  )
```
:::

::: {.column width="50%"}
```{r}
#| label: summarise-grouped
#| code-fold: false
#| code-line-numbers: "3"

gapminder |>
  filter(year == 2007) |>
  group_by(continent) |>
  summarise(
    mean_life = mean(lifeExp),
    n = n()
  )
```
:::

::::

. . .

`group_by()` changes nothing you can see.
It tells the **next** verb to work once per group.
`n()` counts rows, and `count(continent)` is the shortcut for exactly that.

::: notes
This is a data summary table: step 2 of the workflow in one pipeline.
:::

## Practice: summarise {#sec-practice-summarise .smaller}

**Do this now.**

1. Count the countries per continent in 2007 with `count()`.
2. For each continent, compute the median GDP per capita in 2007.
3. **Stretch:** for each year, the mean life expectancy in Europe.
   Did it ever fall?

. . .

```{r}
#| label: practice-summarise
#| code-fold: false

gapminder |>
  filter(year == 2007) |>
  group_by(continent) |>
  summarise(median_gdp = median(gdpPercap)) |>
  arrange(desc(median_gdp))
```

::: notes
Stretch answer: filter Europe, group_by(year), summarise the mean.
It never fell, from 64.4 in 1952 to 77.6 in 2007.
:::

# Your pipeline {#sec-your-pipeline}

## Practice: the whole pipeline {#sec-practice-pipeline .smaller}

**Do this now, in pairs.**
For our nine countries in 2007, compare **EU members** with the rest:

- how many countries in each group
- their mean life expectancy
- their total GDP in billions

Hint: after the join, `eu_member = !is.na(eu_joined)` makes the grouping variable.

::: notes
Six minutes.
Walk the room.
Pairs who finish: the life expectancy gain stretch from the mutate practice.
:::

## The whole pipeline {#sec-pipeline-answer .smaller}

```{r}
#| label: pipeline-answer
#| code-fold: false

gapminder |>
  filter(country %in% lifeexp_wide$country, year == 2007) |> # rows
  left_join(eu_accession, by = "country") |> # add a table
  mutate(
    eu_member = !is.na(eu_joined), # new columns
    gdp_bn = pop * gdpPercap / 1e9
  ) |>
  group_by(eu_member) |> # groups
  summarise(
    n = n(),
    mean_life = mean(lifeExp),
    total_gdp_bn = sum(gdp_bn)
  ) |> # one row per group
  arrange(desc(mean_life)) # sort
```

. . .

Read it aloud as "take gapminder, **and then** ...".
Every line is one step you can run and check on its own.

# Wrap-up {#sec-wrap-up}

## Summary {#sec-summary .smaller}

Today we:

- recapped the **cell options** that decide what a PDF shows
- told **R, Quarto and LaTeX** errors apart
- **installed** and **loaded** packages, and called functions as `package::function()`
- met the **tidyverse** and the three rules of **tidy data**
- **read** a file, made it **tidy** with `pivot_longer()`
- **transformed** it with `filter()`, `select()`, `mutate()` and `arrange()`
- **joined** a second table and **summarised** by group

## Homework {#sec-homework .smaller}

- `hw-04`: wrangle the ParlGov election data with today's verbs, due 7 October.
  Linked from the Schedule.
- A sample solution is published after the due date.

::: {.callout-important}
**Individual feedback is only available through the opt-in path.**
Add an open-source `license.md` to your submission repo to opt in.
:::

::: {.callout-tip}
**Do the DataCamp course "Introduction to the Tidyverse" this week.**
It drills today's verbs until they are automatic, with gapminder as its dataset.
How to join the course group is on the [syllabus](https://qmir-2026-fall.github.io/syllabus.html#sec-practice-materials).
:::

## Next week {#sec-next-week}

**Data visualisation**

- From a table to a figure with `ggplot2`.
- Every plot starts from a tidy table, which is why today came first.

```{r}
#| label: session-info
#| include: false

sessionInfo()
Sys.time() - start_time
```
