---
title: "Week 4: The tidyverse"
---

<!-- Shared format options live in _metadata.yml, so this deck's YAML carries only the title.

     This week teaches code, so the code has to be visible: every teaching chunk opts out of the
     default fold with #| code-fold: false, as in week 03.

     Rhythm: every concept is followed by a "Practice:" slide the room does on the spot, with the
     answer behind a . . . beat. The red line is ONE story: read a messy file, make it tidy, then
     transform, join and summarise, while a single gapminder pipeline grows one verb at a time
     (the new line is highlighted with code-line-numbers). It ends in a pipeline the room writes.

     Time budget (90 min): cell options and errors 4, packages 10, tidyverse and tidy data 10,
     read 8, tidy 6, transform 17, join 9, count and summarise 10, pipeline and bonus 14,
     wrap-up 2.
     If the room runs slow, cut arrange first, then pivot_wider, then the kable bonus.

     data/week04-lifeexp-wide.csv is gapminder's life expectancy for the nine countries of the
     hw-03 `europe` table, pivoted wide (one column per year) and written once with:
       gapminder |> filter(country %in% nine) |> select(country, year, lifeExp) |>
         pivot_wider(names_from = year, values_from = lifeExp) |> write_csv(...)

     data/eu-members.csv is typed by hand: every EU member state with its accession date, plus
     the United Kingdom with its exit date (europa.eu, "EU countries"). Founders date from the
     EEC treaty entering into force, 1958-01-01, and Germany means West Germany until 1990.
     Country names follow gapminder's spelling (Czech Republic, Slovak Republic) so the join is
     clean. Six members are not in gapminder at all (Luxembourg, Cyprus, the Baltics, Malta).

     Students read both files from their raw GitHub URLs, so both must be pushed before class. -->

```{r}
#| label: setup
#| include: false

start_time <- Sys.time()

library(tidyverse) # wrangling
library(here) # project-root-relative paths
library(gapminder) # demo data

# _quarto.yml is a project-root marker, so anchor here() at the repo root.
here::i_am("website/slides/week04.qmd")

# Short tibble printouts, so a result fits under its code on a slide.
options(pillar.print_min = 5, pillar.print_max = 5, width = 70)
```

## Week 4 goals {#sec-week04-goals .smaller}

By the end of today, you will be able to:

1. Install and load packages, and call a function as `package::function()`.
2. Say what the tidyverse is and what makes data tidy.
3. Read a file and reshape it into tidy form.
4. Transform, join and summarise data in one readable pipeline.
5. Answer a question with a pipeline, and report the answer as a captioned table.

## Today {#sec-today .smaller}

1. **Before we start:** cell options and error messages, in brief
2. **Packages:** install, load, `package::function()`
3. **The tidyverse** and tidy data
4. **Read, tidy, transform:** one dataset, one growing pipeline
   - `read_csv()`, `pivot_longer()`, the `dplyr` verbs, `left_join()`, `count()`, `summarise()`
5. **Your pipeline:** from a question to an answer

::: notes
Same deal as last week: type along.
Positron open, a fresh `.qmd` or the console.
:::

## Where today fits {#sec-where-today-fits .smaller}

Today starts **step 2, the data summary**.
Every data summary you will write this term is a pipeline like the one we build today.

{{< include _workflow-8step.qmd >}}

::: notes
One minute.
Point at step 2 only.
:::

# Before we start: cell options and errors {#sec-before-we-start}

## Cell options you have met {#sec-cell-options-table .smaller}

::: notes
Two minutes, no practice.
They used label, echo, eval and error in hw-03.
Point at include, message, warning and the two captions as the new ones.
:::

::: {.small}
| Option | What it does |
|---|---|
| `label: my-name` | Names the cell. Unique, and required. |
| `echo: false` | Runs the code, hides it, shows the result. |
| `eval: false` | Shows the code, does not run it. |
| `include: false` | Runs the code, shows nothing at all. |
| `error: true` | Shows an error and keeps rendering. |
| `message: false` | Hides messages, such as package startup text. |
| `warning: false` | Hides warnings. |
| `fig-cap: "..."` | Captions a figure, with a `fig-` label. |
| `tbl-cap: "..."` | Captions a table, with a `tbl-` label. |

: The cell options this course uses, one `#|` line each at the top of a cell. Set under `execute:` in the YAML header, an option applies to every cell instead. {#tbl-cell-options}
:::

## Errors: which layer is complaining? {#sec-render-relay .smaller}

::: notes
Two minutes, no practice.
The one idea: the first word of the message tells you which runner dropped the baton.
:::

`.qmd` → **R** runs the cells → `.md` → **Quarto and Pandoc** convert → `.tex` → **LaTeX** typesets → `.pdf`

::: {.small}
| Layer | The message looks like | Where to look |
|---|---|---|
| **R** | `Error:` or `Error in f():`, then `Quitting from hw.qmd:6-10 [cell-label]` | The cell named in square brackets, at those lines. |
| **Quarto** | `ERROR: YAMLException: ... (6:20)` | The YAML header, at line 6, column 20. |
| **LaTeX** | `compilation failed- error`, then a line number `l.172` | Your prose: a special character such as `%`, `&`, `$` or a stray `\`. |

: The three layers of a PDF render, how each one reports a failure, and where to go to fix it. {#tbl-error-layers}
:::

::: {.callout-tip}
**Only the first error counts.**
Scroll up to the first message, fix that, and render again.
:::

# Packages {#sec-packages}

## Packages: functions other people wrote {#sec-what-packages .smaller}

::: notes
Callback to week 3: they wrote `celsius_to_fahrenheit()`.
A package is that, times a few hundred, with documentation, shared.
:::

Last week you wrote your own function.
A **package** is a bundle of functions, their help pages, and sometimes data, that someone wrote and shared.

. . .

- R itself ships with the **base** packages: `mean()`, `round()`, `data.frame()`.
- **CRAN**, R's package archive, holds more than 20,000 more, checked and free.
- Nearly every method in this course lives in a package: `dplyr` today, `ggplot2` next week, `brms` later.

## Install once, load every session {#sec-install-load .smaller}

:::: {.columns}

::: {.column width="50%"}
**Install** downloads the package to your computer.
Once, in the **console**.

```{r}
#| label: install-demo
#| code-fold: false
#| eval: false

install.packages("gapminder")
```
:::

::: {.column width="50%"}
**Load** makes its functions available.
Every session, at the top of your `.qmd`.

```{r}
#| label: load-demo
#| code-fold: false

library(gapminder) # demo data
```
:::

::::

. . .

Like an app on your phone: you install it once, and you open it every time you use it.

::: {.callout-important}
**`install.packages()` never goes into a `.qmd`.**
It would reinstall on every render, and Quarto refuses to run it anyway.
:::

## Quarto stops when a package is missing {#sec-missing-package .smaller}

If a `library()` call names a package that is not installed, the render stops at that cell:

```text
Error in `library()`:
! there is no package called 'gapminder'
Quitting from hw-04.qmd:6-10 [setup]
```

. . .

- That is an **R** error, reported through Quarto.
- The fix is in the **console**, not in the document: `install.packages("gapminder")`, then render again.
- It also happens when the name is misspelled: `library(gapmindr)` gives the same message.

::: notes
This is the most common first-render failure on a new laptop.
Everybody who installed a package in a different R version meets it once.
:::

## `package::function()` {#sec-double-colon .smaller}

`::` names the package a function comes from.

```{r}
#| label: double-colon
#| code-fold: false

gapminder::gapminder |> nrow()
```

. . .

Three reasons when it can be useful:

- **Clarity:** the reader sees where a function lives.
- **No `library()` needed** for a one-off call, such as `readxl::read_excel()`.
- **Name clashes:** two packages can both have a `filter()`.
  `library(tidyverse)` tells you which one wins: `✖ dplyr::filter() masks stats::filter()`.

`?dplyr::filter` opens the help page of that exact function.
Most packages also have good online documentation, such as [dplyr.tidyverse.org](https://dplyr.tidyverse.org).

::: {.callout-tip}
**Rule of thumb: more than three calls, load the package.**
Once you use a package's functions more than three times, add a `library()` line to your setup cell instead of repeating `package::` every time.
:::

## Practice: install and load {#sec-practice-packages .smaller}

::: notes
Circulate.
Anyone whose `library(tidyverse)` fails installs it now.
:::

**Do this now.**

1. In the **console**, install `gapminder`.
2. At the top of your `.qmd`, load `tidyverse` and `gapminder` in a setup cell.
   Run the cell and read the startup message.
   Which packages were attached, and what is masked?
3. Type `gapminder` and run it.
   How many rows and columns?
4. Run `?gapminder`.
   What does `gdpPercap` measure?

## Practice: install and load, solution {#sec-practice-packages-solution .smaller}

:::: {.columns}

::: {.column width="55%"}
```{r}
#| label: practice-packages
#| code-fold: false

library(tidyverse) # wrangling
library(gapminder) # demo data

gapminder
```
:::

::: {.column width="45%"}
1. `install.packages("gapminder")`, in the console only.
2. The nine core packages, from `dplyr` to `tidyr`.
   `dplyr::filter()` and `dplyr::lag()` mask their namesakes from `stats`.
3. **1704 rows** and **6 columns**.
4. GDP per capita, in US dollars and adjusted for inflation.
:::

::::

# The tidyverse {#sec-tidyverse}

## One grammar across many packages {#sec-tidyverse-philosophy .smaller}

The **tidyverse** is a family of packages built on one shared philosophy:

. . .

- **Tidy data in, tidy data out:** every function takes a table and returns a table.
- **Data first:** the table is always the first argument, so every step chains with `|>`.
- **Functions are verbs:** `filter()`, `select()` and `summarise()` read like what they do.
- **Written for humans:** consistent names and arguments, so learning one package teaches you the next.

. . .

`library(tidyverse)` loads the core packages in one line.

::: notes
Two minutes.
The payoff of the design: tomorrow's package already feels familiar.
:::

## What is tidy data? {#sec-tidy-data .smaller}

> Tidy datasets are all alike, but every messy dataset is messy in its own way.
>
> Hadley Wickham

. . .

A dataset is **tidy** when:

1. Each **variable** has its own **column**.
2. Each **observation** has its own **row**.
3. Each **value** has its own **cell**.

. . .

This is the data frame from week 3, with one rule added: **decide what a row is**.
In gapminder, a row is one country in one year.

## Practice: is this tidy? {#sec-practice-tidy .smaller}

**Do this now.**
Tidy or not?
If not, which rule does it break?

:::: {.columns}

::: {.column width="50%"}
::: {.small}
| country | year | lifeExp |
|---|---|---|
| Germany | 1952 | 67.5 |
| Germany | 2007 | 79.4 |

: Layout A. {#tbl-layout-a}

| country | 1952 | 2007 |
|---|---|---|
| Germany | 67.5 | 79.4 |
| France | 67.4 | 80.7 |

: Layout B. {#tbl-layout-b}
:::
:::

::: {.column width="50%"}
::: {.small}
| country | year | measure | value |
|---|---|---|---|
| Germany | 2007 | lifeExp | 79.4 |
| Germany | 2007 | pop | 82400996 |

: Layout C. {#tbl-layout-c}

| country | year | life_range |
|---|---|---|
| Germany | 1952-2007 | 67.5-79.4 |

: Layout D. {#tbl-layout-d}
:::
:::

::::

. . .

A is tidy.
B hides the variable `year` in its column names.
C stores two variables in rows.
D puts two values in each cell.

::: notes
Hands up for each one before revealing.
B is the one we meet on the next slides: it is our file.
:::

## The core packages, along the data cycle {#sec-core-packages .smaller}

**Read** → **Tidy** → **Transform** → Visualise → Model → Communicate

::: {.small}
| Stage | Package | For | Key functions |
|---|---|---|---|
| Read | `readr` | Text files such as CSV | `read_csv()`, `write_csv()` |
| Tidy | `tidyr` | Reshaping | `pivot_longer()`, `pivot_wider()` |
| Transform | `dplyr` | The verbs of data manipulation | `filter()`, `mutate()`, `summarise()` |
| Transform | `stringr`, `forcats`, `lubridate` | Text, factors, dates | `str_detect()`, `fct_reorder()`, `ymd()` |
| Visualise | `ggplot2` | Plots, next week | `ggplot()` |
| Underneath | `tibble`, `purrr` | Data frames, repetition | `tibble()`, `map()` |

: The core tidyverse packages, ordered along the data cycle. Today covers the first three stages. {#tbl-core-packages}
:::

::: {.small}
::: {.callout-note}
**In a real project the cycle loops.**
A plot reveals an odd value, and you go back to tidy or transform again, sometimes many times over.
:::
:::

::: notes
Do not read the table.
Point at the three stages we cover, then move on.
:::

# Read {#sec-read}

## Every format has a reader {#sec-import-formats .smaller}

::: {.small}
| Format | Extension | Function | Loaded by `library(tidyverse)`? |
|---|---|---|---|
| Comma-separated | `.csv` | `readr::read_csv()` | yes |
| Semicolon-separated | `.csv` | `readr::read_csv2()` | yes |
| Tab-separated | `.tsv` | `readr::read_tsv()` | yes |
| R's own format | `.rds` | `readr::read_rds()` | yes |
| Excel | `.xlsx`, `.xls` | `readxl::read_excel()` | installed, not loaded |
| Stata | `.dta` | `haven::read_dta()` | installed, not loaded |
| SPSS | `.sav` | `haven::read_sav()` | installed, not loaded |
| JSON | `.json` | `jsonlite::fromJSON()` | installed, not loaded |
| Parquet | `.parquet` | `arrow::read_parquet()` | install it yourself |

: Common data formats and the function that reads each one. Installing the tidyverse also installs `readxl`, `haven` and `jsonlite`, but `library(tidyverse)` loads only `readr`. {#tbl-import-formats}
:::

::: notes
Point out the German Excel row: a CSV that comes out as one column is usually a semicolon file.
The next homework reads an `.xlsx` with `readxl::read_excel()`.
:::

## Reading a file {#sec-read-csv .smaller}

```{r}
#| label: read-csv
#| code-fold: false

path <- here("website", "slides", "data", "week04-lifeexp-wide.csv")
lifeexp_wide <- read_csv(path)
glimpse(lifeexp_wide)
```

. . .

::: {.small}
- **Assign the result.** `read_csv()` alone prints the data and forgets it.
- `here()` builds the path from your project folder, so it works on any computer.
  In your homework repository that is `here("data", "parlgov.xlsx")`.
- `glimpse()` shows every column with its type: the first thing to run on new data.
:::

## Practice: read it yourself {#sec-practice-read .smaller}

**Do this now.**
`read_csv()` also reads straight from the web.
Both files live in the course repository on GitHub.

```{r}
#| label: practice-read
#| code-fold: false
#| eval: false

base <- "https://raw.githubusercontent.com/qmir-2026-fall/qmir-2026-fall.github.io/main/website/slides/data/"
lifeexp_wide <- read_csv(paste0(base, "week04-lifeexp-wide.csv")) # paste0(): paste() without spaces
eu_members <- read_csv(paste0(base, "eu-members.csv"))
```

1. Read both files and `glimpse()` each.
2. Which rule of tidy data does `lifeexp_wide` break?
   A tidy version has one row per country and year: how many rows will it have?
3. What is a row of `eu_members`, and which type did `accession_date` get?

. . .

::: {.small}
1. `glimpse(lifeexp_wide)` and `glimpse(eu_members)`.
2. It is Layout B: `year` hides in the column names.
   Nine countries times twelve years makes **108 rows**.
3. One EU member state, 28 rows including the United Kingdom, which left in 2020.
   `accession_date` is a **`<date>`**: `read_csv()` recognises `2004-05-01` as a date by itself.
:::

# Tidy {#sec-tidy}

## `pivot_longer()`: wide to tidy {#sec-pivot-longer .smaller}

```{r}
#| label: pivot-longer
#| code-fold: false

lifeexp <- lifeexp_wide |>
  pivot_longer(
    cols = -country, # every column except country
    names_to = "year", # the column names become this variable
    values_to = "lifeExp" # the cell values become this one
  )
lifeexp
```

::: notes
Read the call aloud as a sentence: take every column except country, put the names into year and the values into lifeExp.
:::

## `pivot_wider()`: the way back {#sec-pivot-wider .smaller}

Wide is not wrong.
It is how a **table for a reader** usually looks, and `pivot_wider()` gets you there:

```{r}
#| label: pivot-wider
#| code-fold: false

lifeexp |>
  pivot_wider(names_from = year, values_from = lifeExp)
```

. . .

::: {.callout-tip}
**Tidy for the computer, wide for the reader.**
Keep your working data tidy, and pivot wider only at the very end, for a table in your report.
:::

## Practice: pivot it {#sec-practice-pivot .smaller}

::: notes
The transition to gapminder is the key sentence here.
Our file was a slice of gapminder all along.
:::

**Do this now.**

1. Pivot `lifeexp_wide` into a tidy tibble called `lifeexp`.
2. Check with `nrow()` that it has 108 rows.
3. **Stretch:** pivot it back to wide.

## Practice: pivot it, solution {#sec-practice-pivot-solution .smaller}

:::: {.columns}

::: {.column width="50%"}
```{r}
#| label: practice-pivot
#| code-fold: false

lifeexp <- lifeexp_wide |>
  pivot_longer(
    cols = -country,
    names_to = "year",
    values_to = "lifeExp"
  )
nrow(lifeexp)
```
:::

::: {.column width="50%"}
```{r}
#| label: practice-pivot-stretch
#| code-fold: false

# stretch
lifeexp |>
  pivot_wider(
    names_from = year,
    values_from = lifeExp
  )
```
:::

::::

From here on we use the full `gapminder`: it is this tibble for all 142 countries, with three more variables.

# Transform {#sec-transform}

## The `dplyr` verbs {#sec-dplyr-verbs .smaller}

:::: {.columns}

::: {.column width="60%"}
::: {.small}
| Verb | What it does |
|---|---|
| `filter()` | Keep rows that meet a condition |
| `slice()` | Keep rows by position |
| `select()` | Keep or drop columns |
| `mutate()` | Create new columns or change old ones |
| `arrange()` | Sort rows |
| `left_join()` | Add columns from another table |
| `count()` | Count rows per value of a column |
| `group_by()` | Split rows into groups |
| `summarise()` | Summary statistics, such as a mean, per group |

: The core `dplyr` verbs covered today. Each takes a table and returns a table, so they chain into one pipeline. {#tbl-dplyr-verbs}
:::
:::

::: {.column width="40%"}
::: {.small}
Every verb:

- takes a **table first**, so it pipes
- names columns **without quotes** and without `$`
- returns a **new table** and leaves the old one unchanged
:::

We build **one pipeline**, one verb at a time.
The highlighted line is the new one.
:::

::::

## `filter()`: keep rows {#sec-filter .smaller}

```{r}
#| label: pipe-filter
#| code-fold: false
#| code-line-numbers: "2"

gapminder |>
  filter(continent == "Europe", year == 2007)
```

. . .

- A condition is built from the **logical operators** of week 3: `==`, `!=`, `<`, `>`, `%in%` and friends ([the table in week 3](week03.html#/sec-logical)).
- A comma between conditions means **and**, `|` means **or**.

## `select()`: keep columns {#sec-select .smaller}

```{r}
#| label: pipe-select
#| code-fold: false
#| code-line-numbers: "3"

gapminder |>
  filter(continent == "Europe", year == 2007) |>
  select(country, lifeExp, pop, gdpPercap)
```

. . .

A minus drops a column instead: `select(-continent)`.

## Practice: filter and select {#sec-practice-filter-select .smaller}

**Do this now.**

1. Keep the Asian countries in 1952, with only `country` and `lifeExp`.
2. Which countries had a life expectancy above 80 in 2007?
3. **Stretch:** keep only the nine countries of our file, in every year.
   Hint: `%in%` and the `country` column of `lifeexp_wide`.

. . .

:::: {.columns}

::: {.column width="50%"}
```{r}
#| label: practice-filter-select-1
#| code-fold: false

gapminder |>
  filter(continent == "Asia", year == 1952) |>
  select(country, lifeExp)
```
:::

::: {.column width="50%"}
```{r}
#| label: practice-filter-select-2
#| code-fold: false

gapminder |>
  filter(lifeExp > 80, year == 2007) |>
  select(country, lifeExp)
```
:::

::::

## Practice: filter and select, the stretch {#sec-practice-filter-select-stretch .smaller}

```{r}
#| label: practice-filter-select-stretch
#| code-fold: false

gapminder |>
  filter(country %in% lifeexp_wide$country)
```

. . .

Nine countries times twelve years: 108 rows, the same as our tidy `lifeexp`, with the three extra columns of gapminder.

## `mutate()`: new or changed columns {#sec-mutate .smaller}

::: notes
`1e6` is a million, `1e9` a billion.
Several columns in one `mutate()`, separated by commas.
Point at the column type under `country`: `<fct>` became `<chr>`.
:::

```{r}
#| label: pipe-mutate
#| code-fold: false
#| code-line-numbers: "4-8"

gapminder |>
  filter(continent == "Europe", year == 2007) |>
  select(country, lifeExp, pop, gdpPercap) |>
  mutate(
    country = as.character(country),
    pop_m = pop / 1e6,
    gdp_bn = pop * gdpPercap / 1e9
  )
```

. . .

- A **new name** creates a new column: `pop_m`, `gdp_bn`.
- An **existing name** changes that column: `country` was a factor (week 3), and `as.character()` makes it plain text.

## `arrange()`: sort rows {#sec-arrange .smaller}

```{r}
#| label: pipe-arrange
#| code-fold: false
#| code-line-numbers: "9"

gapminder |>
  filter(continent == "Europe", year == 2007) |>
  select(country, lifeExp, pop, gdpPercap) |>
  mutate(
    country = as.character(country),
    pop_m = pop / 1e6,
    gdp_bn = pop * gdpPercap / 1e9
  ) |>
  arrange(desc(gdp_bn))
```

. . .

`arrange()` sorts ascending, and `desc()` flips it.

## Practice: mutate and arrange {#sec-practice-mutate-arrange .smaller}

**Do this now.**

1. For 2007, add a column `pop_m` with the population in millions.
2. Sort so that the most populous country comes first.
   Which five countries lead?
3. **Stretch:** in 2007, which country had the largest economy, its total GDP in billions?
   Is it also the richest per person?

. . .

```{r}
#| label: practice-mutate-arrange
#| code-fold: false

gapminder |>
  filter(year == 2007) |>
  mutate(pop_m = pop / 1e6) |>
  arrange(desc(pop_m)) |>
  select(country, pop_m)
```

## Practice: mutate and arrange, the stretch {#sec-practice-mutate-arrange-stretch .smaller}

:::: {.columns}

::: {.column width="50%"}
**Largest economy**

```{r}
#| label: practice-mutate-arrange-stretch
#| code-fold: false

gapminder |>
  filter(year == 2007) |>
  mutate(gdp_bn = pop * gdpPercap / 1e9) |>
  arrange(desc(gdp_bn)) |>
  select(country, gdp_bn, gdpPercap)
```
:::

::: {.column width="50%"}
**Richest per person**

```{r}
#| label: practice-mutate-arrange-stretch-2
#| code-fold: false

gapminder |>
  filter(year == 2007) |>
  mutate(gdp_bn = pop * gdpPercap / 1e9) |>
  arrange(desc(gdpPercap)) |>
  select(country, gdp_bn, gdpPercap)
```
:::

::::

. . .

The **United States** has the largest economy, and China the second largest, at only about 5,000 dollars per person.
**Norway** is the richest per person, although it ranks only 40th by total GDP.
Total size and wealth per person are different questions.

# Join {#sec-join}

## Four ways to join {#sec-join-types .smaller}

A **join** combines two tables by matching rows on a shared **key** column, such as `country`.

::: {.small}
| Join | Keeps |
|---|---|
| `left_join(x, y)` | Every row of `x`, with the matching columns of `y`, and `NA` where `y` has no match |
| `right_join(x, y)` | Every row of `y`, with the matching columns of `x`, and `NA` where `x` has no match |
| `inner_join(x, y)` | Only the rows that match in **both** tables |
| `full_join(x, y)` | Every row of both tables, and `NA` wherever there is no match |

: The four joins of `dplyr`. They differ only in which rows survive when a key has no partner in the other table. {#tbl-join-types}
:::

. . .

::: {.callout-tip}
**In practice, `left_join()` is nearly always the one you want.**
Your main tibble sits on the left, and the join adds information to it from a second table.
No row of your main data is lost, so it is the only join we practise today.
:::

## `left_join()`: add columns from another table {#sec-left-join .smaller}

::: notes
Before running the join, ask the prediction question on the next slide.
:::

gapminder does not know who is in the EU.
`eu_members` does:

```{r}
#| label: eu-members
#| code-fold: false

eu_members <- read_csv(here("website", "slides", "data", "eu-members.csv"))
```

. . .

```{r}
#| label: pipe-join
#| code-fold: false
#| code-line-numbers: "3"

gapminder |>
  filter(continent == "Europe", year == 2007) |>
  left_join(eu_members, by = "country") |>
  select(country, lifeExp, accession_date)
```

## Practice: predict the join {#sec-practice-join .smaller}

**Do this now.**
Before you run the join, write down:

1. How many rows will the result have?
2. What will `accession_date` hold for Norway?
3. What happens if one table spells `"Czechia"` and the other `"Czech Republic"`?

. . .

1. **30**, the European countries in 2007: `left_join()` keeps every row of the left table.
   Cyprus and Malta are in `eu_members` but not in gapminder, so they do not appear.
2. **`NA`**: Norway has no match, and `NA` is R's missing value from week 3.
3. Czechia silently gets `NA` too.

::: {.callout-warning}
**A join matches text exactly.**
One spelling difference and the row gets `NA`, without any error.
Always check how many `NA`s a join produced.
:::

# Summarise {#sec-summarise}

## `count()`: how many rows per value {#sec-count .smaller}

```{r}
#| label: count
#| code-fold: false

gapminder |>
  filter(year == 2007) |>
  count(continent)
```

. . .

- `count()` returns one row per value of the column, with the number of rows in `n`.
- `count(continent, sort = TRUE)` puts the largest group first.
- Run it on any grouping variable first: how big is each group?

## `summarise()`: summary statistics per group {#sec-summarise-intro .smaller}

::: notes
This is a data summary table: step 2 of the workflow in one pipeline.
:::

:::: {.columns}

::: {.column width="50%"}
```{r}
#| label: summarise-all
#| code-fold: false

gapminder |>
  filter(year == 2007) |>
  summarise(
    mean_life = mean(lifeExp),
    n = n()
  )
```
:::

::: {.column width="50%"}
```{r}
#| label: summarise-grouped
#| code-fold: false
#| code-line-numbers: "3"

gapminder |>
  filter(year == 2007) |>
  group_by(continent) |>
  summarise(
    mean_life = mean(lifeExp),
    n = n()
  )
```
:::

::::

. . .

Without groups, `summarise()` returns one row for the whole table.
`group_by()` changes nothing you can see: it tells `summarise()` to work **once per group**.
`n()` counts the rows of each group.

## Practice: count and summarise {#sec-practice-summarise .smaller}

**Do this now.**

1. Count the countries per continent in 2007 with `count()`.
2. For each continent, compute the median GDP per capita in 2007.
3. **Stretch:** for each year, the mean life expectancy in Europe.
   Did it ever fall?

. . .

:::: {.columns}

::: {.column width="50%"}
```{r}
#| label: practice-summarise-1
#| code-fold: false

gapminder |>
  filter(year == 2007) |>
  count(continent)
```
:::

::: {.column width="50%"}
```{r}
#| label: practice-summarise-2
#| code-fold: false

gapminder |>
  filter(year == 2007) |>
  group_by(continent) |>
  summarise(median_gdp = median(gdpPercap)) |>
  arrange(desc(median_gdp))
```
:::

::::

## Practice: count and summarise, the stretch {#sec-practice-summarise-stretch .smaller}

:::: {.columns}

::: {.column width="55%"}
```{r}
#| label: practice-summarise-stretch
#| code-fold: false

gapminder |>
  filter(continent == "Europe") |>
  group_by(year) |>
  summarise(mean_life = mean(lifeExp)) |>
  print(n = 12)
```
:::

::: {.column width="45%"}
**It never fell.**
Mean life expectancy in Europe rose in every five-year step, from 64.4 years in 1952 to 77.6 in 2007.

`print(n = 12)` shows all twelve rows instead of the first five.
:::

::::

# Your pipeline {#sec-your-pipeline}

## Practice: from a question to an answer {#sec-practice-pipeline .smaller}

::: notes
Eight minutes.
Walk the room.
Push pairs to run the pipeline after every line, not only at the end.
:::

**The question.**
In 2007, did European countries **inside the EU** live longer, and were they richer, than those outside?

**Do this now.**
With `gapminder` and `eu_members` only, build one pipeline that returns one row per group, with the number of countries, the mean life expectancy and the mean GDP per capita.

. . .

Hints:

- After the join, a country is a member in 2007 if it joined in 2007 or earlier: `eu_member = !is.na(accession_date) & year(accession_date) <= 2007`.
  `year()` comes from `lubridate`, loaded with the tidyverse.
- Why not just `!is.na(accession_date)`?
  Croatia is in the file, but joined in 2013.

## The whole pipeline {#sec-pipeline-answer .smaller}

```{r}
#| label: pipeline-answer
#| code-fold: false

eu_2007 <- gapminder |>
  filter(continent == "Europe", year == 2007) |> # rows
  left_join(eu_members, by = "country") |> # add a table
  mutate(
    eu_member = !is.na(accession_date) & year(accession_date) <= 2007
  ) |> # new column
  group_by(eu_member) |> # groups
  summarise(
    n = n(),
    mean_life = mean(lifeExp),
    mean_gdppc = mean(gdpPercap)
  ) # one row per group
eu_2007
```

. . .

::: {.callout-important}
**Pipelines are powerful, so test them at every step.**
Run the pipeline after each new line and look at what it returns: how many rows, which columns, any `NA`s?
A pipeline you have not checked step by step can be wrong without a single error message.
:::

## The answer {#sec-pipeline-conclusion .smaller}

```{r}
#| label: pipeline-numbers
#| include: false

eu_in <- eu_2007 |> filter(eu_member)
eu_out <- eu_2007 |> filter(!eu_member)
gdp_gap <- round(eu_in$mean_gdppc - eu_out$mean_gdppc, -3)
```

**Yes, on both counts, but by very different margins.**

- The `r eu_in$n` EU members lived `r round(eu_in$mean_life - eu_out$mean_life, 1)` years longer on average than the `r eu_out$n` countries outside: `r sprintf("%.1f", eu_in$mean_life)` against `r sprintf("%.1f", eu_out$mean_life)`.
- Their mean GDP per capita was about `r format(gdp_gap, big.mark = ",")` US dollars higher.

. . .

::: {.callout-note}
**A mean hides who is in the group.**
The countries outside mix Norway and Switzerland, among the richest in the data, with Albania and Bosnia and Herzegovina.
And a difference between groups says nothing yet about whether membership caused it: that is step 8, the limitations.
:::

## Bonus: a table for your PDF {#sec-pipeline-kable .smaller}

A printed tibble is for you.
A reader gets a numbered, captioned table: `knitr::kable()` in a cell with a `tbl-` label and a `tbl-cap`.

```{r}
#| label: tbl-eu-2007
#| tbl-cap: "European countries in 2007, split by EU membership at the time. Members lived a little longer on average and had a clearly higher mean GDP per capita (US dollars, adjusted for inflation). Data: gapminder and the course's EU membership file."
#| code-fold: false

eu_2007 |>
  knitr::kable(
    digits = 1,
    col.names = c(
      "EU member",
      "Countries",
      "Mean life expectancy",
      "Mean GDP per capita"
    )
  )
```

In the text, `@tbl-eu-2007` becomes a clickable "Table 1".

# Wrap-up {#sec-wrap-up}

## Summary {#sec-summary .smaller}

Today we:

- recapped the **cell options** and how to tell **R, Quarto and LaTeX** errors apart
- **installed** and **loaded** packages, and called functions as `package::function()`
- met the **tidyverse** and the three rules of **tidy data**
- **read** a file, made it **tidy** with `pivot_longer()`
- **transformed** it with `filter()`, `select()`, `mutate()` and `arrange()`
- **joined** a second table with `left_join()`, then **counted** and **summarised** by group
- answered a question with **one pipeline**, and turned the result into a captioned table

## Homework {#sec-homework .smaller}

- `hw-04`: wrangle the ParlGov election data with today's verbs, due 7 October.
  Linked from the Schedule.
- A sample solution is published after the due date.

::: {.callout-important}
**Individual feedback is only available through the opt-in path.**
Add an open-source `license.md` to your submission repo to opt in.
:::

::: {.callout-tip}
**Do the DataCamp course "Introduction to the Tidyverse" this week.**
It drills today's verbs until they are automatic, with gapminder as its dataset.
How to join the course group is on the [syllabus](https://qmir-2026-fall.github.io/syllabus.html#sec-practice-materials).
:::

## Next week {#sec-next-week}

**Data visualisation**

- From a table to a figure with `ggplot2`.
- Every plot starts from a tidy table, which is why today came first.

```{r}
#| label: session-info
#| include: false

sessionInfo()
Sys.time() - start_time
```
