---
title: "Glossary"
toc: true
toc-depth: 2
---

```{r}
#| label: setup-glossary
#| echo: false
#| message: false
#| warning: false

library(here) # project-root-relative paths
library(yaml) # read course.yml and glossary.yml

# website/_quarto.yml is itself a project-root marker, so a bare here() would stop at
# website/ rather than the repo root. i_am() anchors it explicitly. See STYLE.md section 4.
here::i_am("website/glossary.qmd")
source(here("website", "_course-helpers.R"))

cfg <- read_course()
entries <- yaml::read_yaml(here("website", "glossary.yml"))

# The CLOSED tag vocabulary. An unknown tag stops the render rather than quietly
# adding a fifteenth near-synonym, which is how the previous course iteration ended
# up with 20 ad-hoc slide font sizes. Widen this list deliberately or not at all.
tag_vocab <- c(
  "workflow", "quarto", "git", "r", "data", "probability", "model", "prior",
  "mcmc", "diagnostics", "regression", "multilevel", "interpretation"
)
bad <- setdiff(unique(unlist(lapply(entries, function(e) e$tags))), tag_vocab)
if (length(bad) > 0) {
  stop(
    "glossary.yml uses tags outside the closed vocabulary: ",
    paste(bad, collapse = ", "),
    ". Add them to tag_vocab in glossary.qmd first, or use an existing tag."
  )
}
```

The vocabulary of the course, collected in one place and sorted alphabetically.
Every entry is tagged with the block in which you first meet it, so you can still read this page
in the same order as the [Schedule](schedule.qmd) if you want to.

Nothing here replaces the slides.
A glossary entry is the short version you can check quickly, not the explanation.

::: {.callout-tip}
**Use this page as a lookup.**
If a term on a slide or in a homework is unfamiliar, search this page first.
If it is missing, say so in class, because the gap is probably not only yours.
:::

```{r}
#| label: glossary-body
#| echo: false
#| message: false
#| warning: false
#| results: asis

# -----------------------------------------------------------------------------
# The page is BUILT from website/glossary.yml. Add a term there, not here.
# Emitted as raw markdown so the letter headings are real headings: they land in
# the TOC and can be linked to. Entries stay pandoc definition lists.
# -----------------------------------------------------------------------------

# Sort key: lowercase, with backticks, parentheses and leading punctuation removed,
# so `here()` files under H and `brms` under B rather than under the backtick.
sort_key <- function(term) {
  k <- tolower(term)
  k <- gsub("[`()]", "", k)
  trimws(k)
}

keys <- vapply(entries, function(e) sort_key(e$term), character(1))
entries <- entries[order(keys, method = "radix")]
letters_of <- toupper(substr(sort(keys, method = "radix"), 1, 1))

alphabet <- LETTERS
present <- alphabet %in% letters_of

# --- the A to Z jump bar. Letters with no entries are greyed, not linked. ---
jump <- vapply(seq_along(alphabet), function(i) {
  if (present[i]) {
    sprintf("[%s](#sec-glossary-%s)", alphabet[i], tolower(alphabet[i]))
  } else {
    sprintf("[%s]{.schedule-tbd}", alphabet[i])
  }
}, character(1))

cat("::: {.glossary-jump}\n", paste(jump, collapse = " "), "\n:::\n\n", sep = "")

# --- tag badges: the block tag first, then the topical tags ---
badges <- function(e) {
  block <- block_name_of(cfg, cfg$blocks[[e$block]]$from)
  out <- sprintf("[%s]{.glossary-tag .glossary-block}", block)
  out <- c(out, sprintf("[%s]{.glossary-tag}", e$tags))
  paste(out, collapse = " ")
}

current <- ""
for (e in entries) {
  first <- toupper(substr(sort_key(e$term), 1, 1))
  if (!identical(first, current)) {
    current <- first
    cat("\n## ", first, " {#sec-glossary-", tolower(first), "}\n\n", sep = "")
  }
  cat(e$term, "\n", sep = "")
  # A definition list item: colon, three spaces, then the lines of the definition
  # indented to match. Blank lines inside an entry would end the item, so they go.
  body <- trimws(e$def)
  lines <- strsplit(body, "\n", fixed = TRUE)[[1]]
  lines <- lines[nzchar(trimws(lines))]
  cat(":   ", lines[1], "\n", sep = "")
  for (l in lines[-1]) cat("    ", trimws(l), "\n", sep = "")
  cat("\n    ", badges(e), "\n\n", sep = "")
}
```

<!--
Adding a term.

Entries live in website/glossary.yml, NOT in this file. This page only sorts,
groups and renders them. The header comment in glossary.yml documents the fields
and the closed tag vocabulary, and an unknown tag stops the render.

The letter headings and their {#sec-glossary-x} ids are generated, so the anchors
stay stable as terms are added and the A to Z jump bar keeps working.
-->
