---
title: "Glossary"
toc: true
toc-depth: 2
---

```{r}
#| label: setup-glossary
#| echo: false
#| message: false
#| warning: false

library(here) # project-root-relative paths
library(yaml) # read course.yml and glossary.yml

# website/_quarto.yml is itself a project-root marker, so a bare here() would stop at
# website/ rather than the repo root. i_am() anchors it explicitly. See .claude/r-code.md.
here::i_am("website/glossary.qmd")
source(here("website", "_course-helpers.R"))

cfg <- read_course()
entries <- yaml::read_yaml(here("website", "glossary.yml"))

# The CLOSED tag vocabulary. An unknown tag stops the render rather than quietly
# adding a fifteenth near-synonym, which is how the previous course iteration ended
# up with 20 ad-hoc slide font sizes. Widen this list deliberately or not at all.
tag_vocab <- c(
  "workflow",
  "quarto",
  "git",
  "r",
  "data",
  "probability",
  "model",
  "prior",
  "mcmc",
  "diagnostics",
  "regression",
  "multilevel",
  "interpretation"
)
bad <- setdiff(unique(unlist(lapply(entries, function(e) e$tags))), tag_vocab)
if (length(bad) > 0) {
  stop(
    "glossary.yml uses tags outside the closed vocabulary: ",
    paste(bad, collapse = ", "),
    ". Add them to tag_vocab in glossary.qmd first, or use an existing tag."
  )
}

# Every entry names the week it is first met, and that week must be a taught one.
# The block badge is derived from it, so a wrong week would silently mislabel the term.
week_ok <- vapply(
  entries,
  function(e) {
    is.numeric(e$week) && length(e$week) == 1 &&
      e$week >= 1 && e$week <= cfg$sessions$count
  },
  logical(1)
)
if (!all(week_ok)) {
  stop(
    "glossary.yml entries with a missing or out-of-range week: ",
    paste(vapply(entries[!week_ok], function(e) e$term, character(1)), collapse = ", "),
    ". Set week: to a number from 1 to ", cfg$sessions$count, "."
  )
}

# Every entry gets an anchor, so entries can link to one another and slides can link here.
# The id is derived from the term unless the entry sets id: itself, which is how a renamed
# term keeps its old links working.
slug <- function(term) {
  k <- tolower(term)
  k <- gsub("[`()]", "", k)
  k <- gsub("[^a-z0-9]+", "-", k)
  k <- gsub("^-+|-+$", "", k)
  paste0("gl-", k)
}
for (i in seq_along(entries)) {
  if (is.null(entries[[i]]$id)) entries[[i]]$id <- slug(entries[[i]]$term)
}
ids <- vapply(entries, function(e) e$id, character(1))
if (anyDuplicated(ids)) {
  stop(
    "glossary.yml has duplicate entry ids: ",
    paste(unique(ids[duplicated(ids)]), collapse = ", "),
    ". Rename a term or set an explicit id: on one of them."
  )
}

# A link to an entry that does not exist would fail silently in the browser, so it stops
# the render instead. Checked: every (#gl-...) link in a def, and every id under see:.
linked <- unlist(lapply(entries, function(e) {
  in_def <- regmatches(e$def, gregexpr("\\(#gl-[a-z0-9-]+\\)", e$def))[[1]]
  c(gsub("^\\(#|\\)$", "", in_def), unlist(e$see))
}))
dangling <- setdiff(linked, ids)
if (length(dangling) > 0) {
  stop(
    "glossary.yml links to entries that do not exist: ",
    paste(unique(dangling), collapse = ", "),
    ". Fix the link, or check the target's id."
  )
}
term_of <- setNames(vapply(entries, function(e) e$term, character(1)), ids)
```

The vocabulary of the course, collected in one place and sorted alphabetically.
Every entry is tagged with the week in which you first meet it, and with that week's block, so you can still read this page in the same order as the [Schedule](schedule.qmd) if you want to.
Terms inside a definition link to their own entries, and software entries link to their official documentation.
The page grows with the course: a term is added in the week it is taught.

Nothing here replaces the slides.
A glossary entry is the short version you can check quickly, not the explanation.

::: {.callout-tip}
**Use this page as a lookup.**
If a term on a slide or in a homework is unfamiliar, search this page first.
If it is missing, say so in class, because the gap is probably not only yours.
:::

```{r}
#| label: glossary-body
#| echo: false
#| message: false
#| warning: false
#| results: asis

# -----------------------------------------------------------------------------
# The page is BUILT from website/glossary.yml. Add a term there, not here.
# Emitted as raw markdown so the letter headings are real headings: they land in
# the TOC and can be linked to. Entries stay pandoc definition lists.
# -----------------------------------------------------------------------------

# Sort key: lowercase, with backticks, parentheses and leading punctuation removed,
# so `here()` files under H and `brms` under B rather than under the backtick.
sort_key <- function(term) {
  k <- tolower(term)
  k <- gsub("[`()]", "", k)
  trimws(k)
}

keys <- vapply(entries, function(e) sort_key(e$term), character(1))
entries <- entries[order(keys, method = "radix")]
letters_of <- toupper(substr(sort(keys, method = "radix"), 1, 1))

alphabet <- LETTERS
present <- alphabet %in% letters_of

# --- the A to Z jump bar. Letters with no entries are greyed, not linked. ---
jump <- vapply(
  seq_along(alphabet),
  function(i) {
    if (present[i]) {
      sprintf("[%s](#sec-glossary-%s)", alphabet[i], tolower(alphabet[i]))
    } else {
      sprintf("[%s]{.schedule-tbd}", alphabet[i])
    }
  },
  character(1)
)

cat(
  "::: {.glossary-jump}\n",
  paste(jump, collapse = " "),
  "\n:::\n\n",
  sep = ""
)

# --- tag badges: the week, its block, then the topical tags ---
# Week 1 sits outside the three blocks (course.yml), so it is labelled as orientation.
badges <- function(e) {
  block <- block_name_of(cfg, e$week)
  if (is.na(block)) block <- "Orientation"
  out <- sprintf("[Week %d]{.glossary-tag .glossary-block}", as.integer(e$week))
  out <- c(out, sprintf("[%s]{.glossary-tag .glossary-block}", block))
  out <- c(out, sprintf("[%s]{.glossary-tag}", e$tags))
  paste(out, collapse = " ")
}

current <- ""
for (e in entries) {
  first <- toupper(substr(sort_key(e$term), 1, 1))
  if (!identical(first, current)) {
    current <- first
    cat("\n## ", first, " {#sec-glossary-", tolower(first), "}\n\n", sep = "")
  }
  # The term carries its anchor, so glossary.html#gl-... lands on it.
  cat("[", e$term, "]{#", e$id, "}\n", sep = "")
  # A definition list item: colon, three spaces, then the lines of the definition
  # indented to match. Blank lines inside an entry would end the item, so they go.
  body <- trimws(e$def)
  lines <- strsplit(body, "\n", fixed = TRUE)[[1]]
  lines <- lines[nzchar(trimws(lines))]
  cat(":   ", lines[1], "\n", sep = "")
  for (l in lines[-1]) {
    cat("    ", trimws(l), "\n", sep = "")
  }
  # Further paragraphs of the same item: indented four spaces after a blank line.
  if (length(e$see) > 0) {
    see <- sprintf("[%s](#%s)", term_of[unlist(e$see)], unlist(e$see))
    cat("\n    See also: ", paste(see, collapse = ", "), ".\n", sep = "")
  }
  if (!is.null(e$docs)) {
    cat("\n    Documentation: [", e$docs$label, "](", e$docs$url, ")\n", sep = "")
  }
  cat("\n    ", badges(e), "\n\n", sep = "")
}
```

<!--
Adding a term.

Entries live in website/glossary.yml, NOT in this file. This page only sorts,
groups and renders them. The header comment in glossary.yml documents the fields
and the closed tag vocabulary, and an unknown tag stops the render.

The letter headings and their {#sec-glossary-x} ids are generated, so the anchors
stay stable as terms are added and the A to Z jump bar keeps working. Each entry also
gets its own {#gl-...} anchor, so entries can link to each other with [text](#gl-...).
-->
