---
title: "Schedule"
toc: true
toc-depth: 2
---

```{r}
#| label: setup-schedule
#| echo: false
#| message: false
#| warning: false

library(here) # project-root-relative paths
library(yaml) # read course.yml and the per-homework meta.yml

# website/_quarto.yml is itself a project-root marker, so a bare here() would stop at
# website/ rather than the repo root. i_am() anchors it explicitly. See STYLE.md section 4.
here::i_am("website/schedule.qmd")
source(here("website", "_course-helpers.R"))

cfg <- read_course()
```

Your roadmap for the term.
One session per week, and each row links the slides, the homework starter repo, and, after the due date, the sample solution.

::: {.schedule-legend}
| | |
|---|---|
| {{< fa chalkboard >}} | **Slides.** The session's deck: concepts, theory, and the hands-on part. |
| {{< fa laptop-code >}} | **Homework.** Opens GitHub's new-repository page with the course organisation already selected. |
| {{< fa file-circle-check >}} | **Sample solution.** Published after the due date, for every homework. |
| {{< fa book >}} | **Resources.** Further readings and online learning resources for the week. |
| [{{< fa chalkboard >}}]{.schedule-tbd} | Greyed out means not published yet. |
| [none]{.schedule-none} | There is no homework in that session at all. |
:::

::: {.callout-important}
**Read this before you start a homework.**
Everything below decides whether your work is found and counted.

**There are `r cfg$homework$count` homework assignments, one in every session except the first and the last.**
The first session is orientation and the last is synthesis and the exam briefing, so neither sets work.
Every session from week `r cfg$homework$first_week` to week `r cfg$homework$last_week` has exactly one.
Week 1 and week `r cfg$sessions$count` are marked [none]{.schedule-none} in the tables below.

**Every homework opens the moment its session ends, and is due `r cfg$homework$due_offset_days` days later.**
So a homework assigned in one session is due before the next one starts.
The {{< fa laptop-code >}} link appears as soon as the homework is released, and the {{< fa file-circle-check >}} sample solution appears once the deadline has passed.

**The {{< fa laptop-code >}} link opens GitHub's new-repository page, already filled in from the week's template repo.**
Check two fields before you click "Create repository".

**Owner must stay `qmir-2026-fall`.**
The link pre-selects it, but the dropdown is editable and GitHub's own default is your personal account.
A repo created on your personal account is invisible to the tracker and will not be counted as a submission, no matter what is in it.

**Repository name must be exactly `hw-NN-<username>`,** with your own GitHub username.
The name is pre-filled as `hw-NN-USERNAME`, and `USERNAME` is a placeholder you replace.
The tracker looks the repo up by that exact name.

If the organisation does not appear in the Owner dropdown, you have not accepted your invitation to the course organisation yet.
Check your email, accept it, and reload the page.
:::

The term runs in three blocks.
Week 1 sits outside them, because it shows the whole workflow up front.

```{r}
#| label: schedule-tables
#| echo: false
#| message: false
#| warning: false
#| results: asis

# -----------------------------------------------------------------------------
# These tables are BUILT, not typed. Sources of truth:
#   course.yml                -> term dates, org, sessions, blocks, planned topics
#   website/slides/weekNN.qmd -> the week's topic (its YAML title), which WINS
#   homework/hw-NN/meta.yml   -> homework title + due date
#   website/resources/hw-NN-solution.pdf -> published sample solution
#   website/resources/NN_resources.qmd   -> the week's resources page
# A link appears as soon as its file exists. Nothing to remember to swap on release.
# Dates come from session_dates(), which skips the cancelled dates in course.yml.
# cfg / here() / pad() are set up in the setup-schedule chunk and the helpers file.
# -----------------------------------------------------------------------------
n_wk <- cfg$sessions$count
dates <- session_dates(cfg)
org <- cfg$github$org

tbd <- "[]{.schedule-tbd}--"
# Grey, not a link: planned but not yet published.
planned <- function(x) sprintf("[%s]{.schedule-tbd}", x)
icon <- function(name) sprintf("{{< fa %s >}}", name)
icon_link <- function(name, href) sprintf("[%s](%s)", icon(name), href)
icon_off <- function(name) sprintf("[%s]{.schedule-tbd}", icon(name))

# --- topic: the deck's own YAML title, minus the "Week N: " prefix ---
# The title shape is a contract (STYLE.md section 5.4) and check-authoring.ps1
# enforces it, so this strip is safe. With no deck yet, fall back to the planned
# topic from course.yml, greyed so a plan never looks like a published deck.
deck_of <- function(i) here("website", "slides", paste0("week", pad(i), ".qmd"))

topic_of <- function(i) {
  f <- deck_of(i)
  ttl <- if (file.exists(f)) {
    tryCatch(rmarkdown::yaml_front_matter(f)$title, error = function(e) NULL)
  } else {
    NULL
  }
  if (!is.null(ttl)) {
    return(trimws(sub("^Week\\s*\\d+\\s*:\\s*", "", ttl)))
  }
  w <- week_entry(cfg, i)
  if (is.null(w) || is.null(w$topic)) tbd else planned(w$topic)
}
slides_of <- function(i) {
  if (file.exists(deck_of(i))) {
    icon_link("chalkboard", sprintf("slides/week%s.html", pad(i)))
  } else {
    icon_off("chalkboard")
  }
}

# --- homework: exists on disk => link the template repo, due passed => the solution ---
hw_meta <- function(i) {
  f <- here("homework", paste0("hw-", pad(i)), "meta.yml")
  if (!file.exists(f)) {
    return(NULL)
  }
  tryCatch(yaml::read_yaml(f), error = function(e) NULL)
}
# The link is gated on `released:`, NOT on meta.yml merely existing. A meta.yml
# committed ahead of time would otherwise advertise a distribution repo that
# release-homework.ps1 has not created yet, and students would hit a 404.
# The gate is evaluated at RENDER time, so release-homework.ps1 and publish-site.ps1
# are run together at the end of the session.

# A week outside the homework window has no homework AT ALL, which is a different
# statement from "not published yet". A greyed icon says the second, so say the first
# explicitly. The window is course.yml, never a literal week number here.
has_hw <- function(i) {
  i >= cfg$homework$first_week && i <= cfg$homework$last_week
}
no_hw <- "[none]{.schedule-none}"

hw_of <- function(i) {
  if (!has_hw(i)) {
    return(no_hw)
  }
  m <- hw_meta(i)
  if (is.null(m) || is.null(m$released) || as.Date(m$released) > Sys.Date()) {
    return(icon_off("laptop-code"))
  }
  # The Owner dropdown on GitHub's template page defaults to the student's PERSONAL
  # account, which is how the week 02 repos ended up off the org and invisible to
  # tracking.ps1. These query parameters pre-select the org and ask for a private
  # repo. Owner stays editable, so schedule.qmd also says the rule in prose below.
  # Canonical URL shape: CLAUDE.md section 6. Kept in step with release-homework.ps1.
  icon_link(
    "laptop-code",
    sprintf(
      paste0(
        "https://github.com/new?template_owner=%s&template_name=hw-%s",
        "&owner=%s&name=hw-%s-USERNAME&visibility=private"
      ),
      org,
      pad(i),
      org,
      pad(i)
    )
  )
}
sol_of <- function(i) {
  if (!has_hw(i)) {
    return(no_hw)
  }
  m <- hw_meta(i)
  if (is.null(m) || is.null(m$due)) {
    return(icon_off("file-circle-check"))
  }
  pdf <- here("website", "resources", sprintf("hw-%s-solution.pdf", pad(i)))
  # Due means "before the due-day session", so the solution may link from the due date on,
  # which is also what release-solution.ps1 allows.
  if (as.Date(m$due) > Sys.Date() || !file.exists(pdf)) {
    return(icon_off("file-circle-check"))
  }
  icon_link(
    "file-circle-check",
    sprintf("resources/hw-%s-solution.pdf", pad(i))
  )
}

# --- resources: the week's resources page, NN_resources.qmd (see resources/README.md) ---
res_of <- function(i) {
  f <- here("website", "resources", sprintf("%s_resources.qmd", pad(i)))
  if (file.exists(f)) {
    icon_link("book", sprintf("resources/%s_resources.html", pad(i)))
  } else {
    icon_off("book")
  }
}

row_of <- function(i) {
  sprintf(
    "| %s | %s | %s | %s | %s | %s | %s |",
    pad(i),
    format(dates[i], "%d %b"),
    topic_of(i),
    slides_of(i),
    hw_of(i),
    sol_of(i),
    res_of(i)
  )
}

# A cancelled date, printed where it falls so the gap in the dates is explained.
skip_row <- function(d) {
  sprintf(
    "| [--]{.schedule-tbd} | %s | %s | | | | |",
    format(d, "%d %b"),
    planned("No session")
  )
}

# No Due column: the due date is the same rule every week and is stated once above.
header <- c(
  "| Week | Date | Topic | Slides | HW | Solution | Resources |",
  "|-----:|:-----|:------|:------:|:--:|:--------:|:-------:|"
)

skips <- skipped_sessions(cfg)

emit_table <- function(weeks, caption, id) {
  rows <- vapply(weeks, row_of, character(1))
  # Any cancelled date that falls inside this block's span goes in at its place.
  for (k in seq_len(nrow(skips))) {
    aw <- skips$after_week[k]
    if (aw %in% weeks) {
      at <- which(weeks == aw)
      rows <- append(rows, skip_row(skips$date[k]), after = at)
    }
  }
  cat(c(header, rows), sep = "\n")
  cat("\n\n: ", caption, " {#", id, "}\n\n", sep = "")
}

emit_table(
  1,
  "Week 1, orientation. The whole 8-step workflow is shown up front, so every later week has somewhere to attach itself.",
  "tbl-schedule-week1"
)

for (k in seq_along(cfg$blocks)) {
  b <- cfg$blocks[[k]]
  weeks <- seq(b$from, min(b$to, n_wk))
  cat("#### Block ", block_numeral(k), ": ", b$name, "\n\n", sep = "")
  emit_table(
    weeks,
    sprintf(
      "Block %s of the term, %s. A link appears as soon as its source file exists, a greyed topic is the planned topic from course.yml rather than a published deck, and the sample solution appears once the due date has passed.",
      block_numeral(k),
      tolower(b$name)
    ),
    sprintf("tbl-schedule-block%d", k)
  )
}
```

## Take-home exam {#sec-exam}

::: {.callout-note}
**The exam is a complete 8-step Bayesian analysis.**
You see the target in week 1 and every week builds toward it.

**Window:** `r exam_window(cfg)`.

**Distribution:** the `exam` repo on the course org, the same template flow as the weekly homework.

**Submission:** the state of your repository when the deadline passes is what gets graded.

**What to expect:** a dataset, the background you need, and a research question.
You run the analysis yourself, end to end, and submit a reproducible repository with a rendered PDF.
Detailed grading criteria come with the exam.
:::
