---
title: "Schedule"
---

```{r}
#| label: setup-schedule
#| echo: false
#| message: false
#| warning: false

library(here) # project-root-relative paths
library(yaml) # read course.yml and the per-homework meta.yml

# website/_quarto.yml is itself a project-root marker, so a bare here() would stop at
# website/ rather than the repo root. i_am() anchors it explicitly. See STYLE.md section 4.
here::i_am("website/schedule.qmd")

cfg <- yaml::read_yaml(here("course.yml"))

# A PLACEHOLDER in course.yml prints as an honest "to be announced", never as the
# literal word. See course.yml for the keys that are still unset.
shown <- function(x, fallback = "to be announced") {
  if (is.null(x) || !nzchar(x) || identical(x, "PLACEHOLDER")) fallback else x
}
exam_window <- if (identical(shown(cfg$exam$opens), "to be announced") ||
  identical(shown(cfg$exam$closes), "to be announced")) {
  "to be announced"
} else {
  paste(cfg$exam$opens, "to", cfg$exam$closes)
}
```

Your roadmap for the term.
One session per week, and each row of @tbl-schedule links the slides, the homework starter repo,
and, after the due date, the sample solution.

- Class
  - {{< fa chalkboard >}} Instructor-led session: concepts, theory and foundations.
  - {{< fa code >}} Hands-on exercises with guidance.
- Homework ({{< fa laptop-code >}})
  - Assigned at the end of class.
  - Due `r cfg$homework$due_offset_days` days later, before the next session.
- Additional resources ({{< fa book >}})
  - Further readings and online learning resources for the week.

The term runs in three blocks, shown in the Block column: scientific programming fundamentals
first, then the logic of Bayesian inference, then applied modelling.
Week 1 sits outside them, because it shows the whole workflow up front.

::: {.callout-note}
## Take-home exam
The exam is a complete **8-step Bayesian analysis**.
You see the target in week 1 and every week builds toward it.

**Window:** `r exam_window`.

**Distribution:** the `exam` repo on the course org, the same template flow as the weekly
homework.

**Submission:** the state of your repository when the deadline passes is what gets graded.

**What to expect:** a dataset, the background you need, and a research question.
You run the analysis yourself, end to end, and submit a reproducible repository with a rendered
PDF.
Detailed grading criteria come with the exam.
:::

```{r}
#| label: tbl-schedule
#| tbl-cap: "Weekly schedule, grouped into the term's three blocks. A link appears as soon as its
#|   source file exists on disk. A greyed topic is the planned topic from course.yml, not yet a
#|   published deck, and the sample-solution link appears only once the homework due date has
#|   passed."
#| echo: false
#| message: false
#| warning: false

# ─────────────────────────────────────────────────────────────────────────────
# This table is BUILT, not typed. Sources of truth:
#   course.yml                -> term dates, org, sessions, blocks, planned topics
#   website/slides/weekNN.qmd -> the week's topic (its YAML title), which WINS
#   homework/hw-NN/meta.yml   -> homework title + due date
#   website/resources/hw-NN-solution.pdf -> published sample solution
#   website/resources/NN_resources.qmd   -> the week's reading page
# A link appears as soon as its file exists. Nothing to remember to swap on release.
# cfg / here() are set up in the setup-schedule chunk above.
# ─────────────────────────────────────────────────────────────────────────────
n_wk <- cfg$sessions$count
first <- as.Date(cfg$sessions$first_session)
org <- cfg$github$org

pad <- function(i) sprintf("%02d", i)
tbd <- "[]{.schedule-tbd}--"
# Grey, not a link: planned but not yet published.
planned <- function(x) sprintf("[%s]{.schedule-tbd}", x)

# --- block: which of the three teaching blocks a week belongs to ---
# Named ONCE in course.yml, so the schedule and glossary.qmd cannot drift apart.
# The name prints on the block's first week only, to keep the column readable.
block_of <- function(i) {
  for (b in cfg$blocks) {
    if (i >= b$from && i <= b$to) {
      return(if (i == b$from) planned(b$name) else "")
    }
  }
  "" # week 1 sits outside the blocks
}

# --- topic: the deck's own YAML title, minus the "Week N: " prefix ---
# The title shape is a contract (STYLE.md section 5.4) and check-authoring.ps1
# enforces it, so this strip is safe. With no deck yet, fall back to the planned
# topic from course.yml, greyed so a plan never looks like a published deck.
planned_topic <- function(i) {
  hit <- Filter(function(w) identical(w$week, i), cfg$weeks)
  if (length(hit) == 0 || is.null(hit[[1]]$topic)) NULL else hit[[1]]$topic
}
topic_of <- function(i) {
  f <- here("website", "slides", paste0("week", pad(i), ".qmd"))
  ttl <- if (file.exists(f)) {
    tryCatch(rmarkdown::yaml_front_matter(f)$title, error = function(e) NULL)
  } else {
    NULL
  }
  if (!is.null(ttl)) {
    return(trimws(sub("^Week\\s*\\d+\\s*:\\s*", "", ttl)))
  }
  p <- planned_topic(i)
  if (is.null(p)) tbd else planned(p)
}
slides_of <- function(i) {
  f <- here("website", "slides", paste0("week", pad(i), ".qmd"))
  if (file.exists(f)) sprintf("[slides](slides/week%s.html)", pad(i)) else tbd
}

# --- homework: exists on disk => link the template repo, due passed => the solution ---
hw_meta <- function(i) {
  f <- here("homework", paste0("hw-", pad(i)), "meta.yml")
  if (!file.exists(f)) {
    return(NULL)
  }
  tryCatch(yaml::read_yaml(f), error = function(e) NULL)
}
hw_of <- function(i) {
  m <- hw_meta(i)
  if (is.null(m)) {
    return(tbd)
  }
  due <- if (!is.null(m$due)) format(as.Date(m$due), " (due %d %b)") else ""
  sprintf("[hw-%s](https://github.com/%s/hw-%s/generate)%s", pad(i), org, pad(i), due)
}
sol_of <- function(i) {
  m <- hw_meta(i)
  if (is.null(m) || is.null(m$due)) {
    return(tbd)
  }
  pdf <- here("website", "resources", sprintf("hw-%s-solution.pdf", pad(i)))
  if (as.Date(m$due) >= Sys.Date()) {
    return("[]{.schedule-tbd}after due date")
  }
  if (!file.exists(pdf)) {
    return("[]{.schedule-tbd}soon")
  }
  sprintf("[solution](resources/hw-%s-solution.pdf)", pad(i))
}

# --- resources: the week's reading page, NN_resources.qmd (see resources/README.md) ---
res_of <- function(i) {
  f <- here("website", "resources", sprintf("%s_resources.qmd", pad(i)))
  if (file.exists(f)) sprintf("[reading](resources/%s_resources.html)", pad(i)) else tbd
}

weeks <- seq_len(n_wk)
tab <- data.frame(
  Week = pad(weeks),
  Date = format(first + (weeks - 1) * 7, "%d %b"),
  Block = vapply(weeks, block_of, character(1)),
  Topic = vapply(weeks, topic_of, character(1)),
  Slides = vapply(weeks, slides_of, character(1)),
  Homework = vapply(weeks, hw_of, character(1)),
  Solution = vapply(weeks, sol_of, character(1)),
  Resources = vapply(weeks, res_of, character(1)),
  check.names = FALSE,
  stringsAsFactors = FALSE
)
knitr::kable(tab, align = c("r", "l", "l", "l", "l", "l", "l", "l"))
```

The **Homework** links in @tbl-schedule open GitHub's "create from template" page, which makes
your own `hw-NN-<username>` repo.

::: {.callout-important}
**Name that repo exactly `hw-NN-<username>`.**
The tracker looks for it by name, so anything else will not be counted as a submission.
:::
