---
title: "Schedule"
toc: true
toc-depth: 2
---

```{r}
#| label: setup-schedule
#| echo: false
#| message: false
#| warning: false

library(here) # project-root-relative paths
library(yaml) # read course.yml and the per-homework meta.yml

# website/_quarto.yml is itself a project-root marker, so a bare here() would stop at
# website/ rather than the repo root. i_am() anchors it explicitly. See STYLE.md section 4.
here::i_am("website/schedule.qmd")
source(here("website", "_course-helpers.R"))

cfg <- read_course()
```

Your roadmap for the term.
One session per week, and each row links the slides, the homework starter repo, and, after the
due date, the sample solution.

::: {.schedule-legend}
| | |
|---|---|
| {{< fa chalkboard >}} | **Slides.** The session's deck: concepts, theory, and the hands-on part. |
| {{< fa laptop-code >}} | **Homework.** Opens GitHub's "use this template" page. |
| {{< fa file-circle-check >}} | **Sample solution.** Published after the due date, for every homework. |
| {{< fa book >}} | **Reading.** Further readings and online resources for the week. |
| [{{< fa chalkboard >}}]{.schedule-tbd} | Greyed out means not published yet. |
:::

::: {.callout-note}
**Every homework opens the moment its session ends, and is due `r cfg$homework$due_offset_days` days later.**
So a homework assigned in one session is due before the next one starts.
The {{< fa laptop-code >}} link appears here as soon as the homework is released, and the
{{< fa file-circle-check >}} sample solution appears once the deadline has passed.
:::

The term runs in three blocks.
Week 1 sits outside them, because it shows the whole workflow up front.

```{r}
#| label: schedule-tables
#| echo: false
#| message: false
#| warning: false
#| results: asis

# -----------------------------------------------------------------------------
# These tables are BUILT, not typed. Sources of truth:
#   course.yml                -> term dates, org, sessions, blocks, planned topics
#   website/slides/weekNN.qmd -> the week's topic (its YAML title), which WINS
#   homework/hw-NN/meta.yml   -> homework title + due date
#   website/resources/hw-NN-solution.pdf -> published sample solution
#   website/resources/NN_resources.qmd   -> the week's reading page
# A link appears as soon as its file exists. Nothing to remember to swap on release.
# Dates come from session_dates(), which skips the cancelled dates in course.yml.
# cfg / here() / pad() are set up in the setup-schedule chunk and the helpers file.
# -----------------------------------------------------------------------------
n_wk <- cfg$sessions$count
dates <- session_dates(cfg)
org <- cfg$github$org

tbd <- "[]{.schedule-tbd}--"
# Grey, not a link: planned but not yet published.
planned <- function(x) sprintf("[%s]{.schedule-tbd}", x)
icon <- function(name) sprintf("{{< fa %s >}}", name)
icon_link <- function(name, href) sprintf("[%s](%s)", icon(name), href)
icon_off <- function(name) sprintf("[%s]{.schedule-tbd}", icon(name))

# --- topic: the deck's own YAML title, minus the "Week N: " prefix ---
# The title shape is a contract (STYLE.md section 5.4) and check-authoring.ps1
# enforces it, so this strip is safe. With no deck yet, fall back to the planned
# topic from course.yml, greyed so a plan never looks like a published deck.
deck_of <- function(i) here("website", "slides", paste0("week", pad(i), ".qmd"))

topic_of <- function(i) {
  f <- deck_of(i)
  ttl <- if (file.exists(f)) {
    tryCatch(rmarkdown::yaml_front_matter(f)$title, error = function(e) NULL)
  } else {
    NULL
  }
  if (!is.null(ttl)) {
    return(trimws(sub("^Week\\s*\\d+\\s*:\\s*", "", ttl)))
  }
  w <- week_entry(cfg, i)
  if (is.null(w) || is.null(w$topic)) tbd else planned(w$topic)
}
slides_of <- function(i) {
  if (file.exists(deck_of(i))) {
    icon_link("chalkboard", sprintf("slides/week%s.html", pad(i)))
  } else {
    icon_off("chalkboard")
  }
}

# --- homework: exists on disk => link the template repo, due passed => the solution ---
hw_meta <- function(i) {
  f <- here("homework", paste0("hw-", pad(i)), "meta.yml")
  if (!file.exists(f)) {
    return(NULL)
  }
  tryCatch(yaml::read_yaml(f), error = function(e) NULL)
}
# The link is gated on `released:`, NOT on meta.yml merely existing. A meta.yml
# committed ahead of time would otherwise advertise a distribution repo that
# release-homework.ps1 has not created yet, and students would hit a 404.
# The gate is evaluated at RENDER time, so release-homework.ps1 and publish-site.ps1
# are run together at the end of the session.
hw_of <- function(i) {
  m <- hw_meta(i)
  if (is.null(m) || is.null(m$released) || as.Date(m$released) > Sys.Date()) {
    return(icon_off("laptop-code"))
  }
  icon_link("laptop-code", sprintf("https://github.com/%s/hw-%s/generate", org, pad(i)))
}
sol_of <- function(i) {
  m <- hw_meta(i)
  if (is.null(m) || is.null(m$due)) {
    return(icon_off("file-circle-check"))
  }
  pdf <- here("website", "resources", sprintf("hw-%s-solution.pdf", pad(i)))
  if (as.Date(m$due) >= Sys.Date() || !file.exists(pdf)) {
    return(icon_off("file-circle-check"))
  }
  icon_link("file-circle-check", sprintf("resources/hw-%s-solution.pdf", pad(i)))
}

# --- resources: the week's reading page, NN_resources.qmd (see resources/README.md) ---
res_of <- function(i) {
  f <- here("website", "resources", sprintf("%s_resources.qmd", pad(i)))
  if (file.exists(f)) {
    icon_link("book", sprintf("resources/%s_resources.html", pad(i)))
  } else {
    icon_off("book")
  }
}

row_of <- function(i) {
  sprintf(
    "| %s | %s | %s | %s | %s | %s | %s |",
    pad(i), format(dates[i], "%d %b"), topic_of(i),
    slides_of(i), hw_of(i), sol_of(i), res_of(i)
  )
}

# A cancelled date, printed where it falls so the gap in the dates is explained.
skip_row <- function(d) {
  sprintf(
    "| [--]{.schedule-tbd} | %s | %s | | | | |",
    format(d, "%d %b"),
    planned("No session")
  )
}

# No Due column: the due date is the same rule every week and is stated once above.
header <- c(
  "| Week | Date | Topic | Slides | HW | Solution | Reading |",
  "|-----:|:-----|:------|:------:|:--:|:--------:|:-------:|"
)

skips <- skipped_sessions(cfg)

emit_table <- function(weeks, caption, id) {
  rows <- vapply(weeks, row_of, character(1))
  # Any cancelled date that falls inside this block's span goes in at its place.
  for (k in seq_len(nrow(skips))) {
    aw <- skips$after_week[k]
    if (aw %in% weeks) {
      at <- which(weeks == aw)
      rows <- append(rows, skip_row(skips$date[k]), after = at)
    }
  }
  cat(c(header, rows), sep = "\n")
  cat("\n\n: ", caption, " {#", id, "}\n\n", sep = "")
}

emit_table(1, "Week 1, orientation. The whole 8-step workflow is shown up front, so every later week has somewhere to attach itself.", "tbl-schedule-week1")

for (k in seq_along(cfg$blocks)) {
  b <- cfg$blocks[[k]]
  weeks <- seq(b$from, min(b$to, n_wk))
  cat("#### Block ", block_numeral(k), ": ", b$name, "\n\n", sep = "")
  emit_table(
    weeks,
    sprintf(
      "Block %s of the term, %s. A link appears as soon as its source file exists, a greyed topic is the planned topic from course.yml rather than a published deck, and the sample solution appears once the due date has passed.",
      block_numeral(k), tolower(b$name)
    ),
    sprintf("tbl-schedule-block%d", k)
  )
}
```

The **{{< fa laptop-code >}}** links open GitHub's "create from template" page, which makes your
own `hw-NN-<username>` repo.

::: {.callout-important}
**Name that repo exactly `hw-NN-<username>`.**
The tracker looks for it by name, so anything else will not be counted as a submission.
:::

## Take-home exam {#sec-exam}

::: {.callout-note}
**The exam is a complete 8-step Bayesian analysis.**
You see the target in week 1 and every week builds toward it.

**Window:** `r exam_window(cfg)`.

**Distribution:** the `exam` repo on the course org, the same template flow as the weekly
homework.

**Submission:** the state of your repository when the deadline passes is what gets graded.

**What to expect:** a dataset, the background you need, and a research question.
You run the analysis yourself, end to end, and submit a reproducible repository with a rendered
PDF.
Detailed grading criteria come with the exam.
:::
