---
title: "Syllabus"
toc: true
toc-depth: 3
---

```{r}
#| label: setup-syllabus
#| echo: false
#| message: false
#| warning: false

library(here) # project-root-relative paths
library(yaml) # read course.yml

# website/_quarto.yml is itself a project-root marker, so a bare here() would stop at
# website/ rather than the repo root. i_am() anchors it explicitly. See STYLE.md section 4.
here::i_am("website/syllabus.qmd")
source(here("website", "_course-helpers.R"))

cfg <- read_course()

# Every threshold below is DERIVED from course.yml, never typed. The 2026 spring
# syllabus stated 11 homeworks in prose and then patched itself to 10 in a callout,
# because the numbers lived in the text. They live in one file now.
n_sessions <- cfg$sessions$count
att_frac <- cfg$thresholds$attendance
att_min <- ceiling(att_frac * n_sessions)
# The lower tier of the attendance policy: the case-by-case band runs down to 60%.
att_lo_frac <- 0.60
att_lo <- ceiling(att_lo_frac * n_sessions)

n_hw <- cfg$homework$count
hw_frac <- cfg$thresholds$homework
hw_min <- ceiling(hw_frac * n_hw)
```

## Course description {#sec-course-description}

This course introduces quantitative research methods to study questions from international relations and European politics, using modern tools for scientific programming and reproducible research.
You will learn to analyse and visualise data in **R**, document your work in **Quarto**, and manage your projects with **Git** and **GitHub**.
Together these tools support state-of-the-art workflows for transparent, collaborative and well-organised empirical research.
All work is done in **Positron**, an integrated development environment for data science.

In this course, we will learn the **regression paradigm** as a general framework for modeling patterns in data -- essentially, fitting lines and curves to describe relationships among variables -- and **Bayesian inference** as a framework for quantifying uncertainty and learning efficiently from data.

## Course objectives {#sec-course-objectives}

The overall goal is to give you the practical skills and the methodological confidence to write an empirical bachelor thesis in political science.
You will learn to preprocess data, choose and implement appropriate models for continuous and binary outcomes, compute quantities of interest, and interpret results substantively.

Beyond technical proficiency, the course emphasises sound statistical reasoning, effective data visualisation, and communicating empirical findings clearly to a non-technical audience.
Through weekly assignments and a cumulative take-home exam, you will develop the ability to evaluate statistical claims critically in academic research and in public debate alike.

## Course logistics {#sec-logistics}

### General information {#sec-general-information}

The course meets on `r cfg$sessions$weekday`s, `r shown(cfg$sessions$time)`, in
`r shown(cfg$sessions$room)`.
Each session consists of an instructor-led part on concepts and theory, followed by hands-on
exercises.
Homework is assigned at the end of class and is due `r cfg$homework$due_offset_days` days later,
before the next session.
Readings provide background, theory and context, and are linked per week from the
[Schedule](schedule.qmd).

::: {.callout-tip}
**If you have to miss a session, tell me beforehand.**
Email me before the session and I will set up a Zoom room so you can follow it remotely.
:::

### Prerequisites {#sec-prerequisites}

Two different things go by this name, and it is worth keeping them apart.
The first is an administrative condition on taking the module.
The second is what you need on day one to follow what happens in the room.

**Formally.**
The module requires the Basismodul *Einführung in die Internationalen Beziehungen* and the
Basismodule in *Methoden und Statistik*.
Whether you meet those is a question for the module catalogue and the Studienbüro, not for me.

**In terms of what this course assumes: nothing.**
No programming experience is assumed, and no statistics beyond descriptive measures.
Everything the course needs, it teaches.

A basic grasp of statistics does help, and if you would like to shore that up before or during
the term, @sec-practice-materials points at two free ways to do it.

### Module, credits and workload {#sec-module}

```{r}
#| label: workload-figures
#| echo: false

# Derived from course.yml, never typed. See workload() in _course-helpers.R.
wl <- workload(cfg)
```

- **Module:** `r shown(cfg$course$module)`
- **This course:** `r shown(cfg$course$course_type)`
- **Credits:** `r cfg$course$ects` ECTS, which is `r hrs(wl$total)` hours of total workload
- **Recommended semester:** `r shown(cfg$course$target_semester)`
- **Contact time:** `r hrs(wl$contact)` hours, that is `r n_sessions` sessions of
  `r cfg$sessions$minutes` minutes
- **Self-study:** `r hrs(wl$self_study)` hours, about **`r hrs(wl$per_week)` hours a week** across
  the `r n_sessions` teaching weeks

That self-study figure is not decoration.
It is where the weekly homework, the readings and your exam preparation actually live, and the
course is paced on the assumption that you spend roughly that much time on it.

For comparison, the module catalogue budgets a nominal
`r hrs(cfg$course$nominal_contact_hours)` hours of contact time and
`r hrs(cfg$course$nominal_self_study_hours)` hours of self-study for a `r cfg$course$ects` ECTS
course.
The difference is simply that this course meets for `r cfg$sessions$minutes` minutes over
`r n_sessions` weeks rather than for the nominal figure, so a little more of the same total lands
outside the room.

The authoritative version of all of these is the
[module catalogue](https://www.sowi.uni-mannheim.de/studium/im-studium/pruefungen-und-ordnungen/)
and the
[examination regulations](https://www.sowi.uni-mannheim.de/studium/im-studium/pruefungen-und-ordnungen/)
for the B.A. Political Science, not this page.
Where the two disagree, the regulations win.

### Attendance policy {#sec-attendance}

Active participation (*Mitarbeit*) is a required component of the coursework
(*Studienleistung*), on the basis of section 13b of the examination regulations.
This course consists of `r n_sessions` sessions.
Successful participation is presumed if you attend at least
**`r pct(att_frac)` of sessions, that is at least `r att_min` out of `r n_sessions`**.

If attendance falls below `r pct(att_frac)` but stays at or above `r pct(att_lo_frac)`
(**`r att_lo` to `r att_min - 1` out of `r n_sessions`**), successful participation is not
presumed.
In that case the possibility of fulfilling the participation requirement is assessed case by
case.

If attendance falls below `r pct(att_lo_frac)` (**fewer than `r att_lo` out of
`r n_sessions`**), successful participation is normally considered impossible under the
examination regulations.

If sessions are cancelled during the semester, the absolute number of required sessions is
adjusted accordingly, while the percentage thresholds stay unchanged.

### Grading policy {#sec-grading}

Under the examination regulations, assessment consists of **coursework (*Studienleistung*,
pass or fail)** and a **final examination (*Prüfungsleistung*, graded)**.

Successful completion of all coursework requirements is a prerequisite for admission to the final
take-home exam.

#### Coursework (*Studienleistung*, pass or fail)

Coursework has two components, and both must be passed.
This is the model the module catalogue provides for rather than a local invention: it names
*Mitarbeit* and *Hausaufgaben* among the Studienleistungen for this course.

**1. Weekly homework assignments.**
There are **`r n_hw` homework assignments** over the term, one in every session except the first
and the last.
The first session is orientation and the last is synthesis and the exam briefing, so neither sets
work, which leaves weeks `r cfg$homework$first_week` to `r cfg$homework$last_week`.
They are distributed through GitHub and assessed pass or fail.
To fulfil this requirement you must pass at least **`r pct(hw_frac)` of them, that is `r hw_min`
out of `r n_hw`**.
Homework must show a serious and complete attempt at the tasks.
A submission that is empty or clearly insufficient does not count as passed.

**2. Active participation (*Mitarbeit*).**
Active participation, as described in @sec-attendance, is a required coursework component and is
assessed pass or fail.

### The take-home exam {#sec-exam}

The final examination (*Prüfungsleistung*) is a **take-home exam**, and it determines the whole
course grade.
In the module catalogue it goes by its formal name, a written term paper
(*schriftliche Hausarbeit*).
Those are two names for the same thing.

**Window:** `r exam_window(cfg)`.

It is distributed through the `exam` repository on the course GitHub organisation, the same
template flow as the weekly homework.
The last version pushed to your exam repository before the deadline is what gets graded.
Late submissions are not accepted.

**What you receive.**
A dataset, the theoretical background you need, and a small number of stated hypotheses.
You do not have to find a topic, a question or data of your own.

**What you produce.**
A short empirical research report that tests those hypotheses using the eight-step Bayesian
workflow in @sec-backbone.
The rendered PDF follows the structure of an empirical political science paper: a data and
methods (or research design) section, a results (or analysis) section, a discussion including
limitations, and a brief conclusion.
No introduction and no literature review are required, because the background and hypotheses are
given to you.

- The report may not exceed **`r format(cfg$exam$max_words, big.mark = ",")` words**. That limit
  is set by the module catalogue, not by me. The reference list, the appendices and the statutory
  declaration do not count towards it, which is my clarification rather than the catalogue's.
- Use figures and tables wherever they genuinely support the argument.
- Do **not** print raw code, warnings, messages or unformatted R output in the PDF.
- **Appendices are welcome** for diagnostics, robustness checks and supporting material that is
  not essential to the main text.

**What you submit.**

- `firstname_lastname_exam.qmd` and the rendered `firstname_lastname_exam.pdf`, at the root of
  your exam repository.
- The **whole repository**, which is part of the submission. Any further notebooks or scripts are
  inspected and count towards the grade. Treat it as a research project.
- A **signed statutory declaration** on the final page of the PDF, following the
  [university's template](https://www.sowi.uni-mannheim.de/media/Fakultaeten/sowi/Dokumente/Studiengangsmanagement/Eigenstaendigkeitserklaerung_Statutory_Declaration.pdf).
  An electronic signature is fine. A PDF without one cannot be graded.

**How it is graded.**
On the university scale, 1.0 (very good) to 5.0 (insufficient), across the four dimensions in
@tbl-exam-grading.
The full criterion-by-criterion scheme, which is what you should actually work from, is
distributed together with the exam.

| Dimension | What it asks | Points |
|:---|:---|---:|
| Argumentation | Are your modelling choices reasoned and justified in the PDF, given the data and the hypotheses? | 30 |
| Interpretation | Are your posterior results read correctly and used to evaluate the hypotheses? | 30 |
| Formalities | Does the PDF look and read like a section of a published paper? | 15 |
| Code and repository | Is the workflow operationalised in a reproducible, traceable repository? | 25 |
| **Total** | | **100** |

: The four grading dimensions of the take-home exam and their weights. Argumentation, Interpretation and Formalities are assessed from the rendered PDF alone, so a justification that lives only in a code comment does not count. Code and repository is assessed from the repository as a whole. {#tbl-exam-grading}

::: {.callout-important}
**A submission that does not render from a clean clone cannot reach top marks.**
The single largest criterion in the code and repository dimension is that the PDF rebuilds from
your repository on another machine.
Test that before the deadline, not after.
:::

Every tool and workflow the exam needs is taught during the course.
Session 12, the multilevel primer, is the one exception: it is not examined.

### Registration and withdrawal {#sec-registration}

Registration for the *Prüfungsleistung*, deregistration, and the applicable deadlines are handled
by the **Studienbüro** under the examination regulations, not by me.
Check the dates there and keep to them, because responsibility for meeting them sits with you.

If a good reason, illness in particular, prevents you from completing the exam within the window,
section 30 of the examination regulations applies.
You must submit a **written application to the Studienbüro without delay**, with evidence.
The examination board decides, and the burden of proof is on the applicant.
Tell me as well, but telling me is not a substitute for the formal application.

Section 5 additionally allows the deadlines for coursework and examinations to be extended on a
timely written application, for instance for students with children or with dependants requiring
care.

### Disability and chronic illness {#sec-accommodations}

If a disability or a chronic illness makes it harder for you to take an assessment in the form it
is set, you are entitled to compensation that restores equal opportunity, under section 6 of the
examination regulations.
Applications are made in writing and in good time, and applications on grounds of disability or
chronic illness go to the university's
[representative for students with disabilities or chronic illness](https://www.uni-mannheim.de/en/academics/during-your-studies/studying-with-a-disability-or-chronic-illness/).

::: {.callout-tip}
**Talk to me early, and tell me only what you want to.**
The formal route runs through the representative and the examination board, and I do not need to
know the nature of your condition.
What helps me is knowing early that an adjustment is coming, so it can be built in rather than
improvised.
:::

### Language and software policy {#sec-language-software}

The course is taught entirely in English and is built around R, Quarto and Git or GitHub for data
analysis, reproducible research and project management.
Work with these tools, because the exercises and workflows are designed around them.

For the take-home exam you may use another script-based language such as Python or Stata, but this
is not recommended and no guidance is provided for non-R workflows.
The emphasis is on understanding the tools, practising them, and learning to find solutions
independently.

### Hardware {#sec-hardware}

You need a computer available during class.
A standard laptop is sufficient, and there are no particular requirements for RAM, an external
GPU, or other advanced components.

If you do not own a suitable device, the University of Mannheim IT department offers a
[laptop rental service](https://www.uni-mannheim.de/en/it/students/borrowing-laptops/).
Please make use of it if you need it.

### Software {#sec-software}

Install all required software before the first session, following the separate
[installation guide](installation.qmd).
Doing it early means we can sort out any problems on day one and start the programming work
properly in the following session.

### Additional learning resources {#sec-practice-materials}

Two external platforms are worth your time.
Both are voluntary and neither is part of the requirements to pass, but each fills a gap this
course cannot fill on its own in one session a week.

**[DataCamp](https://www.datacamp.com/groups/shared_links/bfecde68ebfcd9af2db396c17af1267ab74165318bfda32072373d1d1ded1a47).**
Use the link to join the course group, which gives you **six months of premium access**.
DataCamp offers in-depth, self-paced practical training in R, coding, data analytics and AI
topics, so it is the right place to drill something until it is automatic, at whatever speed
suits you.
The introduction to R track is a good starting point.

::: {.callout-warning}
**Register with your `uni-mannheim.de` address.**
Premium access is tied to the university account.
If you sign up with a private address the group link will not grant it, and the fix is to start
again with the right address.
:::

**[Duolingo Math](https://www.duolingo.com/math).**
Freemium, and a good starting point if you would like to relaunch yourself into mathematics
fundamentals or introductory statistics.
No prior statistics is assumed in this course, so this is not a prerequisite.
It is simply there if the maths behind a session feels shakier than you would like.

### Mental health support {#sec-mental-health}

Life and study at the University of Mannheim can sometimes feel complicated, challenging, or
overwhelming.
That is completely normal.
You are not expected to understand or memorise course content on the first attempt.

If you ever feel mentally or emotionally strained, you may find useful advice on the university's
[good mental health and studying page](https://www.uni-mannheim.de/en/academics/studying-the-healthy-way/good-mental-health/).
The [help in times of crisis page](https://www.uni-mannheim.de/en/academics/studying-the-healthy-way/help-in-times-of-crisis/)
gives an overview of the counselling and support services available at the university.

## Homework and feedback policy {#sec-feedback-policy}

> **A sample solution is published for every homework. Individual feedback is NOT provided by
> default.** The only individual feedback available is the **optional AI feedback** you can opt
> into by adding an open-source `license.md` to your submission repo.

**What that means in practice.**
A sample solution is published for every homework after its due date, and it is the standard
against which you should read your own work.
Individual comments on individual submissions are not part of this course by default.

**Opting in to AI feedback.**
If you add an open-source `license.md` to your submission repository, you opt in to automated,
individualised feedback.
An AI reads your submission alongside the sample solution and writes a `FEEDBACK.qmd` and
`FEEDBACK.pdf` back into your repository, organised by the eight workflow steps.
No `license.md` means no processing at all.
This is entirely optional, and opting out costs you nothing beyond the individual feedback
itself.

### Late work policy {#sec-late-work}

Late submissions are not accepted.
If you anticipate any difficulty in meeting a deadline, discuss it with me **before** the
deadline, not after.

## AI policy {#sec-ai-policy}

All submissions are checked for plagiarism, and violations of the code of conduct may result in
failing the course or in expulsion.

### AI is encouraged {#sec-ai-encouraged}

You will use AI tools during this course.
That is a prediction, not a guess, and pretending otherwise would produce a policy nobody follows.
So the question here is not *whether* but *how to make AI use visible and trustworthy*.

::: {.callout-tip}
**Use it, and think of it as a tutor.**
Always available, infinitely patient, and no question is too basic.
Lean on it when you are stuck, to explain a concept, to tidy code, to improve your writing.
You take final responsibility for everything you submit.
:::

Two honest caveats.
First, AI is more reliable on formal logic, programming and well-covered general topics than on
narrow substantive questions in international relations, so treat narrow factual claims
sceptically and check the source.
Second, leaning on it *before* you understand the material does hinder learning:
[research suggests AI helps most those who already understand the topic](https://arxiv.org/abs/2306.03102).
For a fuller explanation of why AI can undermine learning, see
[this piece by Andrew Heiss](https://governancef25.classes.andrewheiss.com/syllabus.html#ai-large-language-models-and-bullshit).
Do the thinking first, then let the tool help.

### Say so in the commit message {#sec-ai-rule}

This course asks for one thing in return, and it is a habit rather than a paperwork exercise.

::: {.callout-important}
**When a commit contains work an AI helped with, say so in the commit message.**
That is the whole rule.
You do not have to separate your work from AI-assisted work into different commits, and you do
not have to submit transcripts.
:::

This is the convention open-source software development already uses, so you are learning a real
practice rather than a classroom one.
The canonical form is a trailer line at the end of the message:

```
Tighten the interpretation of the interaction term

Co-Authored-By: Claude <noreply@anthropic.com>
```

A plain sentence in the message body does the same job, and is just as acceptable:

```
Rewrite the prior justification

Wording tidied up with AI. The choice of prior and the argument for it are mine.
```

Short messages can simply name it: `Fix the join, AI helped debug it` or
`Reformat code with AI`.
Say what the assistance did where that is not obvious, and be proportionate.
A whole section drafted with AI is worth a sentence.
A spell check is worth four words.

### Why this instead of a chat transcript {#sec-ai-why}

A blanket declaration that "generative AI was used to improve the language of this manuscript"
cannot be checked by anyone.
It is a claim about a process that left no trace.
Submitting chat transcripts does not work either, because a transcript cannot be matched to the
passages it actually shaped.

Version control solves this, and you are learning it anyway.
Your repository already records every change, who made it, when, and exactly what changed.
Marking where AI entered turns that ordinary record into the disclosure, produced at the moment
the work happens rather than reconstructed weeks later.
It is honest about how mixed this work really is, because a sentence you drafted and a model
tightened belongs to neither pile, and it rests on the same voluntary integrity that scientific
standards have always rested on.

**Where the line is.**
Asking AI to explain a concept, proofread your draft, clean up code you wrote, or help debug an
error message: all fine, all encouraged, all easy to attribute.
Generating an analysis from scratch and submitting it as your own, or systematically hiding AI
involvement so the record shows none: not fine.
The common thread is the same in both directions.
The history must not misrepresent how the work came about.

::: {.callout-important}
**You do not need to know any of this yet.**
Git, commits and commit messages are taught in session 2, and you will practise this convention
in every homework from then on.
By the time the exam arrives it is a habit rather than a rule to memorise.
Nobody is expected to arrive already knowing how to do it.
:::

::: {.callout-note}
**Two different things, both involving AI.**
The policy above is about AI *you* use while doing your own work, which is disclosed in your
commit messages.
The opt-in feedback described in @sec-feedback-policy is different: it is run by me, on an
explicit `license.md` opt-in, and it comments on work you have already submitted.
Using the feedback pipeline is never something you need to declare.
:::

## The backbone: from research question to empirical answer {#sec-backbone}

Everything in this course builds toward one skill: getting from a question about the world to an
answer you can defend, with its uncertainty attached.
That route has eight steps, shown in @tbl-workflow, and they are the same eight steps every time.
The **take-home exam is exactly this route**, run end to end on a dataset you meet at the end of
the term.

{{< include slides/_workflow-8step.qmd >}}

**Several of the words in that table probably mean nothing to you yet.**
That is expected, and it is the honest starting point.
Estimand, prior predictive check, R-hat: learning what these are, why each step exists, and what
goes wrong when you skip one, is what the term is for.
You see the whole route in session 1, before you can walk any of it, so that every later session
has somewhere to attach itself.
By session 13 you will have run all eight steps yourself, more than once.

The [Glossary](glossary.qmd) holds the short version of every term, and each session's slides
hold the long version.

## Session plan {#sec-session-plan}

What each session covers, and what you should be able to do afterwards.
Dates, slides, homework and readings are on the [Schedule](schedule.qmd).

::: {.callout-note}
**This is a plan, not a contract.**
Pace is adjusted to the group.
If a topic needs more time, it gets more time, and the schedule is updated rather than the plan
silently abandoned.
:::

```{r}
#| label: session-plan
#| echo: false
#| message: false
#| warning: false
#| results: asis

# -----------------------------------------------------------------------------
# BUILT from `weeks:` in course.yml, which is also what schedule.qmd reads, so the
# syllabus and the schedule cannot state different topics for the same session.
# Add or reword a session THERE. Dates come from session_dates(), which skips the
# cancelled dates, so the session numbering here matches the schedule exactly.
# -----------------------------------------------------------------------------
dates <- session_dates(cfg)

for (i in seq_len(cfg$sessions$count)) {
  w <- week_entry(cfg, i)
  if (is.null(w)) {
    next
  }

  cat(
    "### Session ",
    i,
    ": ",
    w$topic,
    " {#sec-session-",
    pad(i),
    "}\n\n",
    sep = ""
  )

  meta <- format(dates[i], "%d %B %Y")
  blk <- block_name_of(cfg, i)
  if (!is.na(blk)) {
    meta <- paste0(meta, " · ", blk)
  }
  steps <- unlist(w$steps)
  if (length(steps) > 0) {
    meta <- paste0(
      meta,
      " · workflow step",
      if (length(steps) > 1) "s" else "",
      " ",
      paste(steps, collapse = ", ")
    )
  }
  cat("::: {.session-meta}\n", meta, "\n:::\n\n", sep = "")

  if (!is.null(w$content)) {
    cat(trimws(w$content), "\n\n", sep = "")
  }

  goals <- unlist(w$goals)
  if (length(goals) > 0) {
    cat("**Afterwards you will be able to:**\n\n")
    cat(paste0("- ", goals, collapse = "\n"), "\n\n", sep = "")
  }
}
```
