---
title: "Syllabus"
toc: true
toc-depth: 3
---

```{r}
#| label: setup-syllabus
#| echo: false
#| message: false
#| warning: false

library(here) # project-root-relative paths
library(yaml) # read course.yml

# website/_quarto.yml is itself a project-root marker, so a bare here() would stop at
# website/ rather than the repo root. i_am() anchors it explicitly. See STYLE.md section 4.
here::i_am("website/syllabus.qmd")
source(here("website", "_course-helpers.R"))

cfg <- read_course()

# Every threshold below is DERIVED from course.yml, never typed. The 2026 spring
# syllabus stated 11 homeworks in prose and then patched itself to 10 in a callout,
# because the numbers lived in the text. They live in one file now.
n_sessions <- cfg$sessions$count
att_frac <- cfg$thresholds$attendance
att_min <- ceiling(att_frac * n_sessions)
# The lower tier of the attendance policy: the case-by-case band runs down to 60%.
att_lo_frac <- 0.60
att_lo <- ceiling(att_lo_frac * n_sessions)

n_hw <- cfg$homework$count
hw_frac <- cfg$thresholds$homework
hw_min <- ceiling(hw_frac * n_hw)
```

## Course description {#sec-course-description}

This course introduces quantitative research methods in international relations and European
integration, using modern tools for scientific programming and reproducible research.
You will learn to analyse and visualise data in R, document your work in Quarto, and manage your
projects with Git and GitHub.
Together these tools support state-of-the-art workflows for transparent, collaborative and
well-organised empirical research.
All work is done in Positron, an integrated development environment for data science.

The course teaches the Bayesian framework as its primary approach to inference.
That is a deliberate choice, not an incidental one, and it shapes how models are built,
criticised and reported throughout.

## Course objectives {#sec-course-objectives}

The overall goal is to give you the practical skills and the methodological confidence to write
an empirical bachelor thesis in political science.
You will learn to preprocess data, choose and implement appropriate models for continuous and
binary outcomes, compute quantities of interest, and interpret results accurately.

Beyond technical proficiency, the course emphasises sound statistical reasoning, effective data
visualisation, and communicating empirical findings clearly to a non-technical audience.
Through weekly assignments and a cumulative take-home exam, you will develop the ability to
evaluate statistical claims critically, in academic research and in public debate alike.

## From research question to empirical answer {#sec-through-line}

Everything in this course builds toward one skill: getting from a question about the world to an
answer you can defend, with its uncertainty attached.
That route has eight steps, shown in @tbl-workflow, and they are the same eight steps every time.
The **take-home exam is exactly this route**, run end to end on a dataset you meet at the end of
the term.

{{< include slides/_workflow-8step.qmd >}}

**Several of the words in that table probably mean nothing to you yet.**
That is expected, and it is the honest starting point.
Estimand, prior predictive check, R-hat: learning what these are, why each step exists, and what
goes wrong when you skip one, is what the term is for.
You see the whole route in session 1, before you can walk any of it, so that every later session
has somewhere to attach itself.
By session 13 you will have run all eight steps yourself, more than once.

The [Glossary](glossary.qmd) holds the short version of every term, and each session's slides
hold the long version.

## Session plan {#sec-session-plan}

What each session covers, and what you should be able to do afterwards.
Dates, topics and links are on the [Schedule](schedule.qmd).

::: {.callout-note}
**This is a plan, not a contract.**
Pace is adjusted to the group.
If a topic needs more time, it gets more time, and the schedule is updated rather than the plan
silently abandoned.
:::

```{r}
#| label: session-plan
#| echo: false
#| message: false
#| warning: false
#| results: asis

# -----------------------------------------------------------------------------
# BUILT from `weeks:` in course.yml, which is also what schedule.qmd reads, so the
# syllabus and the schedule cannot state different topics for the same session.
# Add or reword a session THERE. Dates come from session_dates(), which skips the
# cancelled dates, so the session numbering here matches the schedule exactly.
# -----------------------------------------------------------------------------
dates <- session_dates(cfg)

for (i in seq_len(cfg$sessions$count)) {
  w <- week_entry(cfg, i)
  if (is.null(w)) next

  cat("### Session ", i, ": ", w$topic, " {#sec-session-", pad(i), "}\n\n", sep = "")

  meta <- format(dates[i], "%d %B %Y")
  blk <- block_name_of(cfg, i)
  if (!is.na(blk)) meta <- paste0(meta, " · ", blk)
  steps <- unlist(w$steps)
  if (length(steps) > 0) {
    meta <- paste0(
      meta, " · workflow step", if (length(steps) > 1) "s" else "",
      " ", paste(steps, collapse = ", ")
    )
  }
  cat("::: {.session-meta}\n", meta, "\n:::\n\n", sep = "")

  if (!is.null(w$content)) cat(trimws(w$content), "\n\n", sep = "")

  goals <- unlist(w$goals)
  if (length(goals) > 0) {
    cat("**Afterwards you will be able to:**\n\n")
    cat(paste0("- ", goals, collapse = "\n"), "\n\n", sep = "")
  }
}
```

## Course logistics {#sec-logistics}

### General information {#sec-general-information}

Each session consists of an instructor-led part on concepts and theory, followed by hands-on
exercises.
Homework is assigned at the end of class and is due `r cfg$homework$due_offset_days` days later,
before the next session.
Readings provide background, theory and context, and are linked per week from the
[Schedule](schedule.qmd).

There is **one session per week**, so there is no parallel group to switch into.
If you have to miss a session, the slides and the sample solution for that week are published on
the site, and the [Glossary](glossary.qmd) covers the terminology introduced.

### Attendance policy {#sec-attendance}

Active participation (*Mitarbeit*) is a required component of the coursework
(*Studienleistung*).
This course consists of `r n_sessions` sessions.
Successful participation is presumed if you attend at least
**`r pct(att_frac)` of sessions, that is at least `r att_min` out of `r n_sessions`**.

If attendance falls below `r pct(att_frac)` but stays at or above `r pct(att_lo_frac)`
(**`r att_lo` to `r att_min - 1` out of `r n_sessions`**), successful participation is not
presumed.
In that case the possibility of fulfilling the participation requirement is assessed case by
case.

If attendance falls below `r pct(att_lo_frac)` (**fewer than `r att_lo` out of
`r n_sessions`**), successful participation is normally considered impossible under the
examination regulations.

If sessions are cancelled during the semester, the absolute number of required sessions is
adjusted accordingly, while the percentage thresholds stay unchanged.

### Grading policy {#sec-grading}

Under the examination regulations, assessment consists of **coursework (*Studienleistung*,
pass or fail)** and a **final examination (*Prüfungsleistung*, graded)**.

Successful completion of all coursework requirements is a prerequisite for admission to the final
take-home exam.

#### Coursework (*Studienleistung*, pass or fail)

Coursework has two components, and both must be passed.

**1. Weekly homework assignments.**
There are **`r n_hw` homework assignments** over the term.
They are distributed through GitHub and assessed pass or fail.
To fulfil this requirement you must pass at least **`r pct(hw_frac)` of them, that is `r hw_min`
out of `r n_hw`**.
Homework must show a serious and complete attempt at the tasks.
A submission that is empty or clearly insufficient does not count as passed.

**2. Active participation (*Mitarbeit*).**
Active participation, as described in @sec-attendance, is a required coursework component and is
assessed pass or fail.

#### Final examination (*Prüfungsleistung*, graded)

The final examination is a **take-home exam**.

**Window:** `r exam_window(cfg)`.

The exam is distributed through the `exam` repository on the course GitHub organisation, the same
template flow as the weekly homework.
You receive a dataset and a problem description for a cumulative data analysis project.

You submit a complete project repository, which must include:

- fully documented and reproducible code,
- a rendered PDF presenting your results, structured like the data, methods and analysis sections
  of a research paper,
- a statutory declaration following the
  [university's guidelines](https://www.sowi.uni-mannheim.de/media/Fakultaeten/sowi/Dokumente/Studiengangsmanagement/Eigenstaendigkeitserklaerung_Statutory_Declaration.pdf).

The state of the repository on GitHub at the moment the deadline passes is what gets evaluated.

The take-home exam is graded on the university scale, 1.0 (very good) to 5.0 (insufficient), and
it determines the overall course grade.
Every tool and workflow needed to complete it is taught during the course.

### Language and software policy {#sec-language-software}

The course is taught entirely in English and is built around R, Quarto and Git or GitHub for data
analysis, reproducible research and project management.
Work with these tools, because the exercises and workflows are designed around them.

For the take-home exam you may use another script-based language such as Python or Stata, but this
is not recommended and no guidance is provided for non-R workflows.
The emphasis is on understanding the tools, practising them, and learning to find solutions
independently.

### Hardware {#sec-hardware}

You need a computer available during class.
A standard laptop is sufficient, and there are no particular requirements for RAM, an external
GPU, or other advanced components.

If you do not own a suitable device, the University of Mannheim IT department offers a
[laptop rental service](https://www.uni-mannheim.de/en/it/students/borrowing-laptops/).
Please make use of it if you need it.

### Software {#sec-software}

Install all required software before the first session, following the separate
[installation guide](installation.qmd).
Doing it early means we can sort out any problems on day one and start the programming work
properly in the following session.

### Additional learning resources {#sec-practice-materials}

Two external platforms are worth your time.
Both are voluntary and neither is part of the requirements to pass, but each fills a gap this
course cannot fill on its own in one session a week.

**[DataCamp](https://www.datacamp.com/groups/shared_links/bfecde68ebfcd9af2db396c17af1267ab74165318bfda32072373d1d1ded1a47).**
Use the link to join the course group, which gives you **six months of premium access**.
DataCamp offers in-depth, self-paced practical training in R, coding, data analytics and AI
topics, so it is the right place to drill something until it is automatic, at whatever speed
suits you.
The introduction to R track is a good starting point.

**[Duolingo Math](https://www.duolingo.com/math).**
Freemium, and a good starting point if you would like to relaunch yourself into mathematics
fundamentals or introductory statistics.
No prior statistics is assumed in this course, so this is not a prerequisite.
It is simply there if the maths behind a session feels shakier than you would like.

## Homework and feedback policy {#sec-feedback-policy}

> **A sample solution is published for every homework. Individual feedback is NOT provided by
> default.** The only individual feedback available is the **optional AI feedback** you can opt
> into by adding an open-source `license.md` to your submission repo.

**What that means in practice.**
A sample solution is published for every homework after its due date, and it is the standard
against which you should read your own work.
Individual comments on individual submissions are not part of this course by default.

**Opting in to AI feedback.**
If you add an open-source `license.md` to your submission repository, you opt in to automated,
individualised feedback.
An AI reads your submission alongside the sample solution and writes a `FEEDBACK.qmd` and
`FEEDBACK.pdf` back into your repository, organised by the eight workflow steps.
No `license.md` means no processing at all.
This is entirely optional, and opting out costs you nothing beyond the individual feedback
itself.

### Late work policy {#sec-late-work}

Late submissions are not accepted.
If you anticipate any difficulty in meeting a deadline, discuss it with me **before** the
deadline, not after.

### AI policy {#sec-ai-policy}

All submissions are checked for plagiarism, and violations of the code of conduct may result in
failing the course or in expulsion.

#### AI is encouraged {#sec-ai-encouraged}

You will use AI tools during this course.
That is a prediction, not a guess, and pretending otherwise would produce a policy nobody follows.
So the question here is not *whether* but *how to make AI use visible and trustworthy*.

::: {.callout-tip}
**Use it, and think of it as a tutor.**
Always available, infinitely patient, and no question is too basic.
Lean on it when you are stuck, to explain a concept, to tidy code, to improve your writing.
You take final responsibility for everything you submit.
:::

Two honest caveats.
First, AI is more reliable on formal logic, programming and well-covered general topics than on
narrow substantive questions in international relations, so treat narrow factual claims
sceptically and check the source.
Second, leaning on it *before* you understand the material does hinder learning:
[research suggests AI helps most those who already understand the topic](https://arxiv.org/abs/2306.03102).
For a fuller explanation of why AI can undermine learning, see
[this piece by Andrew Heiss](https://governancef25.classes.andrewheiss.com/syllabus.html#ai-large-language-models-and-bullshit).
Do the thinking first, then let the tool help.

#### The one rule: say so in the commit message {#sec-ai-rule}

This course asks for exactly one thing in return, and it is a habit rather than a paperwork
exercise.

::: {.callout-important}
**Your own work first, committed. Then the AI-assisted improvement, committed separately, with a
message that names what the AI did.**
Two commits: one yours, one AI-assisted and labelled as such.
That is the whole policy.
:::

In practice:

1. Write your own draft, whether that is a paragraph, a chunk of code, or a piece of analysis.
   Commit it, with a message describing **your** work.
2. If you then use AI to improve it, whether spell checking, rewording, tidying code or
   restructuring, commit that improved version **separately**, and say in the message what the AI
   did.

Example messages for the AI step:

```
AI spell check
AI wording improvement
AI code formatting
AI suggested restructuring of section 3
```

If your tool offers it, the open-source convention of a trailer line works too, and is what much
of the software world now does:

```
Tighten the interpretation of the interaction term

Co-Authored-By: Claude <noreply@anthropic.com>
```

#### Why this instead of a chat transcript {#sec-ai-why}

A blanket declaration that "generative AI was used to improve the language of this manuscript"
cannot be checked by anyone.
It is a claim about a process that left no trace.
Submitting chat transcripts does not work either, because a transcript cannot be matched to the
passages it actually shaped.

Version control solves this, and you are learning it anyway.
Your repository already records every change, who made it, when, and exactly what changed.
Marking where AI entered turns that ordinary record into the disclosure, produced at the moment
the work happens rather than reconstructed weeks later.
It separates your work from AI-assisted work in the history, it lets me see the trajectory and
judge accordingly, and it rests on the same voluntary integrity that scientific standards have
always rested on.

**Where the line is.**
Asking AI to explain a concept, proofread your draft, clean up code you wrote, or help debug an
error message: all fine, all encouraged, all easy to attribute.
Generating an analysis from scratch and submitting it as your own, or editing AI output before
committing so the AI step never appears: not fine.
The common thread is the same in both directions.
The history must not misrepresent who did the work.

::: {.callout-important}
**You do not need to know any of this yet.**
Git, commits and commit messages are taught in session 2, and you will practise this convention
in every homework from then on.
By the time the exam arrives it is a habit rather than a rule to memorise.
Nobody is expected to arrive already knowing how to do it.
:::

::: {.callout-note}
**Two different things, both involving AI.**
The policy above is about AI *you* use while doing your own work, which is disclosed in your
commit messages.
The opt-in feedback described in @sec-feedback-policy is different: it is run by me, on an
explicit `license.md` opt-in, and it comments on work you have already submitted.
Using the feedback pipeline is never something you need to declare.
:::

### Mental health support {#sec-mental-health}

Life and study at the University of Mannheim can sometimes feel complicated, challenging, or
overwhelming.
That is completely normal.
You are not expected to understand or memorise course content on the first attempt.

If you ever feel mentally or emotionally strained, you may find useful advice on the university's
[good mental health and studying page](https://www.uni-mannheim.de/en/academics/studying-the-healthy-way/good-mental-health/).
The [help in times of crisis page](https://www.uni-mannheim.de/en/academics/studying-the-healthy-way/help-in-times-of-crisis/)
gives an overview of the counselling and support services available at the university.
