---
title: "Homework 03: Practicing the Reproducible Workflow"
author: "Your Name"
date: today
format:
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
---

# Introduction to Key Tools

**Quarto** is an open-source scientific and technical publishing system. It lets you combine
*prose*, code, and output (like tables and figures) in a single document -- and render it
to PDF, HTML, or Word. Think of it as a notebook where your analysis and writing live together.

**R** is a programming language designed for statistical computing and data analysis. It is
widely used in political science, economics, and other social sciences for tasks ranging from
basic summary statistics to complex modelling and visualisation.

**Git** is a version control system that tracks changes to files over time. It lets you record
a history of edits, revert mistakes, and collaborate with others without overwriting each
other's work.

**GitHub** is a web platform that hosts Git repositories online. It makes it easy to share
code, collaborate on projects, and submit assignments -- your instructor can simply look at
your repository to see your work and its full history.



# Exercise 4 -- Practicing R: Data Types, Structures, and Subsetting

## 4.1 -- Vectors and Data Types

Below I create three vectors related to European Parliament election results. Each vector
holds a different data type: country names (character), vote share percentages (numeric),
and whether a party is part of the ruling coalition (logical).

```{r}
# Character vector: country names
countries <- c("Germany", "France", "Italy", "Spain", "Poland")
print(countries)
class(countries)

# Numeric vector: largest party vote share (%) in each country
vote_share <- c(28.9, 22.4, 26.0, 33.1, 36.8)
print(vote_share)
class(vote_share)

# Logical vector: whether the country's government is centre-right led
centre_right <- c(FALSE, FALSE, TRUE, FALSE, TRUE)
print(centre_right)
class(centre_right)
```

Each `class()` call confirms the expected type: `"character"`, `"numeric"`, and `"logical"`
respectively.

Now let's see what happens when we mix types in a single vector:

```{r}
# Mixing types forces "coercion"
mixed <- c("Germany", 28.9, TRUE)
print(mixed)
class(mixed)
```

When R encounters different types in one vector it **coerces** everything to the most
flexible type -- here, `character`. The number `28.9` becomes the string `"28.9"` and
`TRUE` becomes `"TRUE"`. This is called *implicit coercion* and is worth watching out for,
because arithmetic on such a vector will no longer work.



## 4.2 -- Building a Data Frame

Using the three vectors above (plus a fourth column for EU membership), I build a small
data frame with 5 rows and 4 columns.

```{r}
eu_elections <- data.frame(
  country = countries,
  vote_share = vote_share,
  centre_right = centre_right,
  eu_member = c(TRUE, TRUE, TRUE, TRUE, TRUE),
  stringsAsFactors = FALSE
)

# Dimensions
nrow(eu_elections)
ncol(eu_elections)

# Structure
str(eu_elections)

# First few rows
head(eu_elections)
```

`nrow()` returns 5 and `ncol()` returns 4, confirming the shape of the data frame.
`str()` shows the column names, their types, and a preview of values -- a quick sanity
check before any analysis. `head()` prints the first six rows (or fewer if the data frame
is smaller), which is useful for visually confirming everything looks right.



## 4.3 -- Accessing and Subsetting Data

```{r}
# 1. Extract the vote_share column using $
eu_elections$vote_share

# 2. Access a single cell: row 3 (Italy), column 2 (vote_share)
eu_elections[3, 2]

# 3. Filter rows where vote share exceeds 30 %
high_share <- eu_elections[eu_elections$vote_share > 30, ]
high_share
```

The `$` operator pulls out a column as a plain vector, which is handy for quick
calculations. The `[row, column]` notation pinpoints any individual value -- here the
vote share for Italy (26.0). The logical filter `eu_elections$vote_share > 30` creates a
`TRUE`/`FALSE` vector of the same length as the data frame; passing it inside `[ , ]`
keeps only the rows where the condition is `TRUE`, returning Spain and Poland as the two
countries where the largest party won more than 30 % of the vote.



# Reflection

The most confusing part of this workflow was keeping the **render–commit loop** in mind
continuously rather than writing everything first and only then checking. It is tempting
to treat rendering as a final step, but catching errors early (especially in code chunks)
saves a lot of time. On the R side, implicit type coercion -- where mixing types silently
converts values -- was a subtle surprise. The fact that R gives no warning when it turns a
number into a string means you can end up with broken calculations if you are not careful
to check `class()` on vectors you have modified. Building the habit of calling `str()` on
a data frame before doing anything else helps catch these issues early.
