---
title: "Homework 04: Wrangling ParlGov election data"
author: "Your name"
date: last-modified
date-format: MMMM D, YYYY
format:
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| include: false

start_time <- Sys.time()

library(tidyverse) # wrangling
library(here) # project-root-relative paths
```

# Read and explore {#sec-read}

This homework uses **ParlGov**, a database of parties, elections and cabinets in 37 democracies.
The file `data/parlgov.xlsx` is an Excel workbook with several sheets.
`readxl` was installed with the tidyverse but is not loaded by it, so call its functions as `readxl::read_excel()`, as in class.

1. List the sheets of the workbook with `readxl::excel_sheets()`.
   Build the path with `here("data", "parlgov.xlsx")`.
2. Read the `election` sheet into an object called `elections` and the `party` sheet into `parties`.
   The `sheet =` argument of `readxl::read_excel()` picks the sheet.
3. `glimpse()` both.

```{r}
#| label: ex-read
```

4. Answer in prose: how many rows does each table have, and what is one row of each?
5. Which type did `election_date` get?
   In class, `read_csv()` read `accession_date` as a `<date>`.
   Why do you think the two differ?

6. `count()` the values of `election_type`.
   What are the two types, and which is more common?
7. The `variable` sheet documents every column.
   Read it and find out what `left_right` measures and on which scale.

```{r}
#| label: ex-explore
```

# Tidy {#sec-tidy}

1. Is `elections` tidy?
   Say in one sentence what a row is, and check it against the three rules of tidy data.

2. `parties` holds three positions of every party, each on a scale from 0 to 10, in three columns: `left_right`, `state_market` and `liberty_authority`.
   For a question such as "how do the party families differ on each scale", the scale is itself a variable.
   **Before running anything**, predict how many rows a tibble with one row per party and scale will have.
3. Pivot the three columns longer, with the names going to `scale` and the values to `position`, and assign the result to `positions`.
   Check your prediction with `nrow()`.
4. Compute the mean `position` per `scale` with one `group_by()` and `summarise()`.
   The first attempt returns `NA`: you met the fix in homework 03.
5. **Stretch:** pivot `positions` back to wide, and check that you get the columns you started from.

```{r}
#| label: ex-tidy
```

# Transform {#sec-transform}

From here on, work with **national parliament elections** only.

1. Filter `elections` to the rows with `election_type == "parliament"` and assign the result to `parl`.
   How many rows remain?
2. In the same pipeline, add three columns with one `mutate()`:
   - `year`, the year of the election.
     `election_date` is text, so turn it into a date first: `year(as.Date(election_date))`.
   - `seat_share`, the percentage of seats a party won: `seats / seats_total * 100`.
   - `disproportionality`, how far a party's seat share lies from its vote share, in either direction: `abs(seat_share - vote_share)`.
     `abs()` returns the absolute value, so a party 5 points over and a party 5 points under both count as 5.

```{r}
#| label: ex-transform-parl
```

3. Show the most recent German election in `parl`, with only the columns `party_name_english`, `vote_share`, `seats` and `seat_share`, sorted by vote share with the largest party first.
   Hint: find the date first, then `filter()` on it.

```{r}
#| label: ex-transform-germany
```

4. In one or two sentences: which parties won more seat share than vote share, and which won no seats at all despite winning votes?
   What rule of the German electoral system do you think explains the second group?

5. **Stretch:** find the five largest seat shares in the whole of `parl`, with the country, the date, the party, its vote share and its seat share.
   What do the countries have in common?

```{r}
#| label: ex-transform-stretch
```

# Join {#sec-join}

## Party families {#sec-join-families}

`parl` does not know which party family a party belongs to.
`parties` does, in `family_name`, and both tables share the key `party_id`.

1. **Before running anything**, predict how many rows `parl` will have after a `left_join()` that adds `family_name` from `parties`.
2. Join `parties |> select(party_id, family_name)` onto `parl` by `party_id`, and assign the result to `parl_family`.
   Check your prediction, and count the `NA`s in `family_name`.
3. Why select the two columns before the join?
   Try the join once with the whole of `parties` and look at the column names.

```{r}
#| label: ex-join-families
```

## EU membership {#sec-join-eu}

The EU membership file from class also lives on GitHub.
Read it straight from the web:

```{r}
#| label: eu-members-read

base <- "https://raw.githubusercontent.com/qmir-2026-fall/qmir-2026-fall.github.io/main/website/slides/data/"
eu_members <- read_csv(paste0(base, "eu-members.csv"))
```

The key has a different name in each table: `country_name` in ParlGov and `country` in `eu_members`.
`by = join_by(country_name == country)` tells `left_join()` that the two columns are the same key.

1. Build a table of ParlGov's countries with one row each: `parl |> count(country_name)`.
   Then join `eu_members` onto it.
2. **Before running the join**, predict how many rows the result has.
3. How many countries get `NA` in `accession_date`?
   `filter()` them with `is.na(accession_date)` and read the list.
   One of them is an EU member state.
   Which one, and why did the join miss it?

```{r}
#| label: ex-join-eu
```

4. Fix the spelling in `eu_members` so that it matches ParlGov, and overwrite `eu_members` with the fixed version.
   Hint: `mutate(country = if_else(country == "...", "...", country))` replaces one value and keeps every other.
5. Run the join again.
   Show with `count()` that 28 countries now have an accession date and 9 do not.

```{r}
#| label: ex-join-eu-fix
```

::: {.callout-warning}
**A join matches text exactly.**
A spelling difference gives `NA` without any error or warning.
Count the `NA`s after every join, and read the list of rows that failed to match.
:::

# Count and summarise {#sec-summarise}

1. Count the rows of `parl_family` per `family_name`, largest first.
   Two of the values are not party families at all.
   Which ones, and what do they mean?

```{r}
#| label: ex-summarise-count
```

2. For every country, compute the mean `disproportionality` and the number of rows with `n()`.
   Sort so that the most disproportional country comes first.
   `vote_share` has missing values, so remember `na.rm = TRUE`.

```{r}
#| label: ex-summarise-disprop
```

3. In two or three sentences: which countries top the list, and which close it?
   What do you know about how these countries turn votes into seats?

4. **Stretch:** is the largest party getting smaller over time?
   For each election, find the vote share of the largest party with `max()`, then compute the mean of those maxima per decade.
   This needs two rounds of `group_by()` and `summarise()`: first per election, then per decade.
   Hint: `decade = year %/% 10 * 10`, and drop the rows without a vote share before the first round.

```{r}
#| label: ex-summarise-stretch
```

# Your pipeline: a data summary {#sec-pipeline}

::: {.callout-important}
**This section is step 2 of the workflow, the data summary.**
A data summary answers a precise question with a table a reader can understand without your code.
From now on, every analysis you write starts with one.
:::

**The question.**
In national parliament elections since 2000, did the average party of each family win a larger vote share in EU member states than outside?

1. Write **one pipeline**, starting from `parl_family`, that:
   - keeps the elections from 2000 onwards
   - joins the fixed `eu_members`
   - adds `eu_member`, which is `TRUE` if the country was a member **on the day of the election**: `!is.na(accession_date) & accession_date <= as.Date(election_date)`
   - drops the two values of `family_name` that are not families
   - returns one row per family and membership status, with the mean vote share in a column `mean_vote`

   Assign the result to `eu_families`, and run the pipeline after every new line, as in class.

```{r}
#| label: ex-pipeline
```

2. Turn `eu_families` into a table for a reader: `pivot_wider()` so that each family is one row, with one column for members and one for non-members.
   Then pass it to `knitr::kable()`, with `digits = 1` and readable column names.
   The cell below already carries the label `tbl-eu-families`.
   Replace its placeholder `tbl-cap` with a caption a reader can understand without the text: what is in the table, for which elections, and in which unit.

```{r}
#| label: tbl-eu-families
#| tbl-cap: "Replace this with your caption."
```

3. Below the table, write three to four sentences that answer the question.
   - Reference the table with `@tbl-eu-families`.
   - Put at least one number into a sentence with inline code, so that it updates when the data does.
   - Name one thing the mean hides, or one reason why a gap between members and non-members is not yet an effect of EU membership.

::: {.callout-tip}
**Tidy for the computer, wide for the reader.**
Keep `eu_families` tidy, and pivot wider only in the cell that prints the table.
:::

# Reflection {#sec-reflection .unnumbered}

Which step of this homework took you longest, and why?
Is there a verb or a join you are still unsure about?

# Session info and execution time {#sec-session .unnumbered}

```{r}
#| label: session-info

sessionInfo()
```

```{r}
#| label: exec-time

end_time <- Sys.time()
cat(
  "Total execution time:",
  round(difftime(end_time, start_time, units = "mins"), 1),
  "minutes\n"
)
```
